---
title: "Test new grid-based optimization for nu_0"
author: "Dat Do"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
date: "2026-06-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 0. Aim

One of the problem I have been largely ignored is the choosing of rank $k$ in the low-rank part $F$ of the low-rank-plus-diag representation $R_q = F F^\top + D$. Recall the `covariance part` of the loss function:

$$\text{KL}(\text{IW}(\nu_q R_q, \nu_q + J + 1), \text{IW}(\nu_0 R_0 + \delta D_0, \nu_0 + \delta + J + 1)). $$
Choosing rank $k$ as large as $\text{rank}(R_0)$ ($\leq \min\{N_0, J\}$) can help this part achieving 0 loss, because $\nu_0 R_0 + \delta D_0$ is a low-rank-plus-diag matrix with rank of the low-rank part being $\text{rank}(R_0)$. 

1) In all of the previous experiments, $N_0 = 200$ and $J = 1000$. We choose $k = 200$ and it works fine.

2) I was surprised that when $N_0$ is as large as 1000, the KL gap between $\text{IW}(\nu_q R_q, \nu_q + J + 1)$ with $k=200$ and $\text{IW}(\nu_0 R_0 + \delta D_0, \nu_0 + \delta + J + 1))$ is so huge that it might actually dominate the `fit part` in the loss function, therefore sending the optimized $\nu_0$ to very small. 


## 1. Simulate data from a coalescent model

```{r simulate, fig.width=8, fig.height=6}
library(susielove)

seed  <- 1
T_split <- 25
chrom <- "chr22"
start <- 20e6
end   <- 21e6
N     <- 25000L
N0    <- 1000L
J     <- 1000L
h2    <- 0.01

res <- sim_2pop(chrom, start, end,
                T_split = T_split,
                n_gwas = N,
                n_ref = N0,
                J = J,
                h2 = h2,
                seed = seed)

v               <- res$v
R               <- res$R
R0              <- res$R0
beta_true       <- res$beta_true
causal_SNPs_loc <- res$causal_SNPs_loc

plot(v, pch = 16)
points(causal_SNPs_loc, v[causal_SNPs_loc], pch = 16, col = "red")
```

## 2. Try fitting susieR with in-sample and out-of-sample covariance matrices

```{r susie-insample, fig.width=8, fig.height=6}
library(susieR)

fit <- susie_suff_stat(
  XtX = N * R,
  Xty = N * v,
  n = N,
  yty = N
)
susie_plot(fit, y = "PIP", b = beta_true, main = "In-sample cov mat")
```

```{r susie-outsample, fig.width=8, fig.height=6}
library(susieR)

fit_R0 <- susie_suff_stat(
  XtX = N * R0,
  Xty = N * v,
  n = N,
  yty = N
)
susie_plot(fit_R0, y = "PIP", b = beta_true, main = "Out-of-sample cov mat")
```


```{r susie-outsample-diag, fig.width=8, fig.height=6}
d0_inv_sqrt <- 1 / sqrt(diag(R0))
LD0 <- diag(d0_inv_sqrt) %*% R0 %*% diag(d0_inv_sqrt)
d_sqrt <- sqrt(diag(R))
R_tilde <- diag(d_sqrt) %*% LD0 %*% diag(d_sqrt)

fit <- susie_suff_stat(
  XtX = N * R_tilde,
  Xty = N * v,
  n = N,
  yty = N
)
susie_plot(fit, y = "PIP", b = beta_true, main = "Out-of-sample cov mat (corrected diagonal)")
```


## 3. Fit susie_XL with new log-space optimization (nu_0_init = 1000)

```{r prep-inputs}
d0_inv_sqrt <- 1 / sqrt(diag(R0))
LD0 <- diag(d0_inv_sqrt) %*% R0 %*% diag(d0_inv_sqrt)
d_sqrt <- sqrt(diag(R))
R_tilde <- diag(d_sqrt) %*% LD0 %*% diag(d_sqrt)

diag_R <- diag(R)
R_out  <- R_tilde
k      <- 200
delta  <- 1
```

```{r susie-XL-1000, fig.width=8, fig.height=6}
nu_0_init <- 1000
max_iter  <- 30

XL_res_1000 <- susie_XL(v, R_out, diag_R,
                         delta,
                         N, L = 5,
                         k,
                         nu_0_init     = nu_0_init,
                         max_iter_init = 100,
                         step_susie    = 20,
                         step_R        = 100,
                         estimate_nu_0 = TRUE,
                         estimate_delta = FALSE,
                         max_iter      = max_iter,
                         tol           = 1e-4)

b_res <- XL_res_1000$b_res
R_q   <- XL_res_1000$R_q

susie_plot_alpha(
  alpha = b_res$alpha,
  R     = cov2cor(R_q),
  b     = beta_true,
  main  = paste0("SuSiE-XL (nu_0_init=1000) | nu_0=",
                 round(XL_res_1000$nu_0, 1),
                 " | nu_q=", round(XL_res_1000$nu_q, 1))
)
```


**Observation:** Compare this to the file test-new-optimization-delta-ver (when $N_0 = 200$): The optimized $\nu_0$ was non-trivial in the previous file, and it just converges to the lower bound 1 in this case $N_0 = 1000$...

But why is this the case??

