---
title: "Try a new algorithmic approach"
author: "dodat-stats"
date: "2026-06-04"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# 1. Aim

We investigate to see if using a **very mismatched** R0 leads to small $\nu_0$.

SPOIL: It actually does not lead to a small $\nu_0$. Ironically, the regularization strength $\lambda$ matters so much more than the discrepancy between $v$ and out-of-sample $R$ in deciding the optimal $\nu_0$... We have had automatic way to decide $\lambda$, but it only depends on $N_0$. Therefore in this case, where $R0$ is way different than $R$, but because $N_0$ is the same, $\lambda$ is the same and leads to quite large $\nu_0$, which leads to more borrowing of information than we like, and leads to more false positive. 

From a bright side, the current implementation is good when $R_0$ is not too mismatched though (mismatch but not super far away from $R$ (like divergence from a few hundred years apart), like in the previous workflow file). 

Anyway, let's see it in action. We generate R from Asian population and R0 from European population.

```{r, include=FALSE}
library(susielove)
library(susieR)

maximize_log_pdf_null <- function(R0, v, N) {
  J <- length(v)
  
  R0 <- (R0 + t(R0)) / 2
  eig <- eigen(R0, symmetric = TRUE)
  
  # Clamp any floating-point negative eigenvalues to exactly 0
  d <- pmax(eig$values, 0) 
  v_tilde <- as.numeric(crossprod(eig$vectors, v))
  
  # Optimize over log_nu0 to keep parameter scales identical
  neg_log_pdf <- function(params) {
    lambda <- params[1]
    nu0 <- exp(params[2]) # Decode log(nu0)
    
    d_lambda <- (1 - lambda) * d + lambda
    
    log_det <- sum(log(d_lambda))
    qf <- N * sum((v_tilde^2) / d_lambda)
    
    inner_log <- 1 + (1 / nu0) * qf
    
    term1 <- lgamma((nu0 + J + 2) / 2)
    term2 <- lgamma((nu0 + 2) / 2)
    term3 <- (J / 2) * log(nu0 * pi)
    term4 <- 0.5 * log_det
    term5 <- ((nu0 + J + 2) / 2) * log(inner_log)
    
    log_lik <- term1 - term2 - term3 - term4 - term5
    
    # Scale objective by J to prevent gradient collapse in finite-differencing
    return(-log_lik / J)
  }
  
  init_lambda <- 0.5
  init_log_nu0 <- log(5.0) 
  
  epsilon <- 1e-5 
  
  # nlminb is significantly more robust for this topology
  opt_res <- nlminb(
    start = c(init_lambda, init_log_nu0),
    objective = neg_log_pdf,
    lower = c(epsilon, log(epsilon)),
    upper = c(1 - epsilon, log(1e5))
  )
  
  return(list(
    lambda = opt_res$par[1],
    nu_0 = exp(opt_res$par[2]),
    # Scale the objective back up by J to get true log likelihood
    max_log_pdf = -opt_res$objective * J,
    convergence = opt_res$convergence, # 0 means successful
    message = opt_res$message
  ))
}

grid_search_single_effect <- function(R_col1, j, R0,
                                      lambda_range = c(1e-4, 0.2),
                                      nu0_range = c(10, 1000),
                                      n_grid = 50) {
  p <- length(R_col1)
  I_p <- diag(p)

  r_11 <- R_col1[j]
  R_21 <- R_col1[-j]

  lambda_grid <- seq(lambda_range[1], lambda_range[2], length.out = n_grid)
  nu0_grid    <- seq(nu0_range[1], nu0_range[2], length.out = n_grid)

  grid <- expand.grid(lambda = lambda_grid, nu_0 = nu0_grid)
  grid$neg_log_lik <- NA_real_

  for (i in seq_len(nrow(grid))) {
    lambda <- grid$lambda[i]
    nu_0   <- grid$nu_0[i]

    Rbar <- (1 - lambda) * R0 + lambda * I_p

    rbar_11 <- Rbar[j, j]
    Rbar_21 <- Rbar[-j, j, drop = FALSE]
    Rbar_22 <- Rbar[-j, -j, drop = FALSE]

    Rbar_22.1 <- Rbar_22 - (Rbar_21 %*% t(Rbar_21)) / rbar_11
    log_det_Rbar_22.1 <- as.numeric(determinant(Rbar_22.1, logarithm = TRUE)$modulus)

    mu <- Rbar_21 * (r_11 / rbar_11)
    diff <- R_21 - mu
    inv_Rbar_22.1 <- solve(Rbar_22.1)
    Q <- as.numeric(t(diff) %*% inv_Rbar_22.1 %*% diff)

    alpha <- (nu_0 + 2) / 2
    beta  <- (nu_0 * rbar_11) / 2

    ll_r11 <- alpha * log(beta) - lgamma(alpha) -
      (alpha + 1) * log(r_11) - (beta / r_11)

    term1 <- lgamma((nu_0 + p + 2) / 2) - lgamma((nu_0 + 3) / 2)
    term2 <- - ((p - 1) / 2) * log(pi)
    term3 <- - ((p - 1) / 2) * log(r_11 / rbar_11)
    term4 <- - 0.5 * log_det_Rbar_22.1
    term5 <- - ((nu_0 + p + 2) / 2) * log(1 + (rbar_11 / r_11) * Q)

    ll_R21_given_r11 <- term1 + term2 + term3 + term4 + term5

    grid$neg_log_lik[i] <- -(ll_r11 + ll_R21_given_r11)
  }

  best <- which.min(grid$neg_log_lik)

  return(list(
    lambda_opt  = grid$lambda[best],
    nu_0_opt    = grid$nu_0[best],
    neg_log_lik = grid$neg_log_lik[best],
    grid        = grid
  ))
}

optimize_nu0_single_effect <- function(R_col1, j, R0, lambda, init_nu0 = 50) {

  p <- length(R_col1)
  I_p <- diag(p)

  r_11 <- R_col1[j]
  R_21 <- R_col1[-j]

  Rbar <- (1 - lambda) * R0 + lambda * I_p

  rbar_11 <- Rbar[j, j]
  Rbar_21 <- Rbar[-j, j, drop = FALSE]
  Rbar_22 <- Rbar[-j, -j, drop = FALSE]

  Rbar_22.1 <- Rbar_22 - (Rbar_21 %*% t(Rbar_21)) / rbar_11
  inv_Rbar_22.1 <- solve(Rbar_22.1)
  log_det_Rbar_22.1 <- as.numeric(determinant(Rbar_22.1, logarithm = TRUE)$modulus)

  mu <- Rbar_21 * (r_11 / rbar_11)
  diff <- R_21 - mu
  Q <- as.numeric(t(diff) %*% inv_Rbar_22.1 %*% diff)

  neg_log_lik <- function(nu_0) {
    alpha <- (nu_0 + 2) / 2
    beta  <- (nu_0 * rbar_11) / 2

    ll_r11 <- alpha * log(beta) - lgamma(alpha) -
      (alpha + 1) * log(r_11) - (beta / r_11)

    term1 <- lgamma((nu_0 + p + 2) / 2) - lgamma((nu_0 + 3) / 2)
    term2 <- - ((p - 1) / 2) * log(pi)
    term3 <- - ((p - 1) / 2) * log(r_11 / rbar_11)
    term4 <- - 0.5 * log_det_Rbar_22.1
    term5 <- - ((nu_0 + p + 2) / 2) * log(1 + (rbar_11 / r_11) * Q)

    ll_R21_given_r11 <- term1 + term2 + term3 + term4 + term5

    return(-(ll_r11 + ll_R21_given_r11))
  }

  opt_result <- optim(
    par       = init_nu0,
    fn        = neg_log_lik,
    method    = "L-BFGS-B",
    lower     = 1e-4,
    upper     = 1e6,
    control   = list(maxit = 10000)
  )

  return(list(
    nu_0_opt    = opt_result$par,
    lambda      = lambda,
    neg_log_lik = opt_result$value,
    convergence = opt_result$convergence,
    message     = opt_result$message
  ))
}

optimize_params_single_effect <- function(R_col1, j, R0, init_nu0 = 50, init_lambda = 0.5) {
  
  # Length of the column vector (p dimension)
  p <- length(R_col1)
  
  # Identity matrix of dimension p
  I_p <- diag(p)
  
  # Partition the observed first column of R
  r_11 <- R_col1[j]
  R_21 <- R_col1[-j]
  
  # ---------------------------------------------------------
  # Objective Function (Negative Log-Likelihood)
  # ---------------------------------------------------------
  neg_log_lik <- function(params) {
    nu_0   <- params[1]
    lambda <- params[2]
    
    # 1. Dynamically construct Rbar based on the current lambda
    Rbar <- (1 - lambda) * R0 + lambda * I_p
    
    # 2. Partition the dynamic Rbar matrix
    rbar_11 <- Rbar[j, j]
    Rbar_21 <- Rbar[-j, j, drop = FALSE] # Keep as a column matrix
    Rbar_22 <- Rbar[-j, -j, drop = FALSE]
    
    # 3. Compute the Schur Complement of Rbar dynamically
    # Rbar_22.1 = Rbar_22 - Rbar_21 * (1/rbar_11) * Rbar_12
    Rbar_22.1 <- Rbar_22 - (Rbar_21 %*% t(Rbar_21)) / rbar_11
    
    # Compute inverse and log-determinant (safely)
    inv_Rbar_22.1 <- solve(Rbar_22.1)
    log_det_Rbar_22.1 <- as.numeric(determinant(Rbar_22.1, logarithm = TRUE)$modulus)
    
    # --- A. Log-Likelihood of r_11 (Inverse-Gamma) ---
    alpha <- (nu_0 + 2) / 2
    beta  <- (nu_0 * rbar_11) / 2
    
    ll_r11 <- alpha * log(beta) - lgamma(alpha) - 
      (alpha + 1) * log(r_11) - (beta / r_11)
    
    # --- B. Log-Likelihood of R_21 | r_11 (Multivariate t) ---
    mu <- Rbar_21 * (r_11 / rbar_11)
    diff <- R_21 - mu
    
    # Quadratic form: (R_21 - mu)^T %*% Rbar_22.1^{-1} %*% (R_21 - mu)
    Q <- as.numeric(t(diff) %*% inv_Rbar_22.1 %*% diff)
    
    term1 <- lgamma((nu_0 + p + 2) / 2) - lgamma((nu_0 + 3) / 2)
    term2 <- - ((p - 1) / 2) * log(pi)
    term3 <- - ((p - 1) / 2) * log(r_11 / rbar_11)
    term4 <- - 0.5 * log_det_Rbar_22.1
    term5 <- - ((nu_0 + p + 2) / 2) * log(1 + (rbar_11 / r_11) * Q)
    
    ll_R21_given_r11 <- term1 + term2 + term3 + term4 + term5
    
    # Return total negative log-likelihood
    return(-(ll_r11 + ll_R21_given_r11))
  }
  
  # ---------------------------------------------------------
  # Execution of the Optimizer
  # ---------------------------------------------------------
  # "L-BFGS-B" allows us to set both lower and upper bounds
  # nu_0 must be > 0. lambda must be in [0, 1].
  opt_result <- optim(
    par       = c(init_nu0, init_lambda),
    fn        = neg_log_lik,
    method    = "L-BFGS-B",
    lower     = c(1e-4, 1e-4),    # Keep strictly > 0
    upper     = c(Inf, 0.9999),   # lambda caps at 1
    control   = list(maxit = 10000)
  )
  
  return(list(
    nu_0_opt    = opt_result$par[1],
    lambda_opt  = opt_result$par[2],
    neg_log_lik = opt_result$value,
    convergence = opt_result$convergence,
    message     = opt_result$message
  ))
}

```


# 2. An example

## 2.1. Generate silly R0

```{r}
sim_asn_eur <- function(chrom, start, end, n_gwas, n_ref, J, h2, seed) {

  stdpopsim <- reticulate::import("stdpopsim")
  msprime   <- reticulate::import("msprime")

  set.seed(seed)

  # 1. Get Realistic Genetic Map from stdpopsim
  species <- stdpopsim$get_species("HomSap")
  contig  <- species$get_contig(chrom, left = as.integer(start), right = as.integer(end))

  # 2. Get OutOfAfrica_3G09 Demography (YRI, CEU, CHB)
  demo_model <- species$get_demographic_model("OutOfAfrica_3G09")
  demography <- demo_model$model

  # 3. Run Simulation
  ts <- msprime$sim_ancestry(
    samples = reticulate::dict(CHB = as.integer(n_gwas), CEU = as.integer(n_ref)),
    demography = demography,
    recombination_rate = contig$recombination_map,
    random_seed = as.integer(seed)
  )

  ts <- msprime$sim_mutations(ts, rate = 1.29e-8, random_seed = as.integer(seed))

  # 4. Extract Genotypes
  chb_id <- NULL
  ceu_id <- NULL
  for (p in reticulate::iterate(ts$populations())) {
    if (!is.null(p$metadata) && p$metadata$name == "CHB") chb_id <- p$id
    if (!is.null(p$metadata) && p$metadata$name == "CEU") ceu_id <- p$id
  }

  samples_gwas <- ts$samples(population = as.integer(chb_id))
  samples_ref  <- ts$samples(population = as.integer(ceu_id))

  G_all <- ts$genotype_matrix()

  to_diploid <- function(hap_idx) {
    idx <- as.integer(hap_idx) + 1
    g_hap <- G_all[, idx, drop = FALSE]
    g_dip <- g_hap[, seq(1, ncol(g_hap), by = 2)] + g_hap[, seq(2, ncol(g_hap), by = 2)]
    return(t(g_dip))
  }

  G_gwas <- to_diploid(samples_gwas)
  G_ref  <- to_diploid(samples_ref)

  # 5. Filter for Common Variants (MAF > 5%)
  maf_gwas <- colMeans(G_gwas) / 2
  maf_ref  <- colMeans(G_ref) / 2

  common_mask <- (maf_gwas > 0.05) & (maf_gwas < 0.95) & (maf_ref > 0.05) & (maf_ref < 0.95)
  n_common <- sum(common_mask)
  if (n_common < J) stop(paste0("Only ", n_common, " common variants found but J = ", J, ". Try a larger region."))
  G_gwas <- G_gwas[, common_mask, drop = FALSE]
  G_ref  <- G_ref[, common_mask, drop = FALSE]

  J_all <- min(ncol(G_gwas), J)
  max_start <- J_all - J
  start_idx <- if (max_start > 0) sample(0:max_start, 1) else 0
  cols_to_keep <- (start_idx + 1):(start_idx + J)
  G  <- G_gwas[, cols_to_keep, drop = FALSE]
  G0 <- G_ref[, cols_to_keep, drop = FALSE]

  # 6. Generate summary stats
  X <- sweep(G, 2, colMeans(G), "-")
  N <- nrow(X)

  G0_centered <- sweep(G0, 2, colMeans(G0), "-")

  R0 <- stats::cov(G0_centered)
  R0[is.na(R0)] <- 0

  R <- stats::cov(X)
  R[is.na(R)] <- 0

  num_causal <- 2
  first_causal_SNP <- sample(1:J, 1)
  second_causal_SNP <- which.min(R[first_causal_SNP, ]^2)
  causal_SNPs_loc <- c(first_causal_SNP, second_causal_SNP)

  beta_true <- rep(0, J)
  beta_true[causal_SNPs_loc] <- stats::runif(num_causal) + 0.01

  var_expl <- sum((X %*% beta_true)^2 / N)
  scale <- h2 / var_expl
  beta_true <- beta_true * sqrt(scale)

  y <- (X %*% beta_true) + sqrt(1 - h2) * stats::rnorm(N)
  y_std <- sqrt(sum((y - mean(y))^2) / N)
  y <- (y - mean(y)) / y_std

  v <- (t(X) %*% y) / N

  return(list(
    v = as.vector(v),
    R = R,
    R0 = R0,
    beta_true = beta_true,
    causal_SNPs_loc = causal_SNPs_loc
  ))
}

```

```{r}
seed = 2
chrom <- "chr22"
start <- 20e6
end   <- 21e6
N     <- 25000L
N0 = 500L
J     <- 500L
h2    <- 0.01

res <- sim_asn_eur(chrom, start, end,
                n_gwas = N,
                n_ref = N0,
                J = J,
                h2 = h2,
                seed = seed)

v               <- res$v
R               <- res$R
R0              <- res$R0
beta_true       <- res$beta_true
causal_SNPs_loc <- res$causal_SNPs_loc

## standardized
d = 1 / sqrt(diag(R))
v = d * v
R = cov2cor(R)
R0 = cov2cor(R0)

plot(v, pch = 16)
points(causal_SNPs_loc, v[causal_SNPs_loc], pch = 16, col = "red")
```



## 2.2 Visualize two correlation matrices

Sorry the color scheme is a bit hard to see but if we squeeze our eyes a bit then we can see that the two matrices are totally different. 

```{r, fig.width=8, fig.height=6}
par(mfrow = c(1, 2))

image(R0[1:100, 1:100], main = "Out-of-sample (EUR)", col = heat.colors(15))
image(R[1:100, 1:100], main = "In-sample corr (CHB)", col = heat.colors(15))
par(mfrow = c(1, 1))

```

## 2.3 Fit susie with in-sample R

```{r, fig.width=8, fig.height=6}
fit_R <- susie_suff_stat(
  XtX = N * R,
  Xty = N * v,
  n = N,
  yty = N,
  L = 5
)
susie_plot(fit_R, y = "PIP", b = beta_true, main = "In-sample cov mat")
```

## 2.4 Fit susie with silly R

```{r, fig.width=8, fig.height=6}
fit_R_silly <- susie_suff_stat(
  XtX = N * R0,
  Xty = N * v,
  n = N,
  yty = N,
  L = 5,
  residual_variance = 1
)
susie_plot(fit_R_silly, y = "PIP", b = beta_true, main = "Silly corr mat")
```
Slightly regularized

```{r, fig.width=8, fig.height=6}
R_tilde = (1 - 1 / N0) * R0 + 1 / N0 * diag(J)
fit_R0 <- susie_suff_stat(
  XtX = N * R0,
  Xty = N * v,
  n = N,
  yty = N,
  L = 5
)
susie_plot(fit_R_silly, y = "PIP", b = beta_true, main = "Silly corr mat")

print(summary(fit_R0)$cs)
```

## 2.5 Fit susielove with silly R

```{r}
# R_out  <- compute_wen_stephens(R_0 = R_tilde, N_ref = N0)
# R_out = diag(J)
R_out = R_tilde
k      <- min(N0, J)
diag_R = diag(R)

love_res_WS <- susie_love_lambda(v, R_out, diag_R,
                              lambda = 0,
                              N, L = 5,
                              k,
                              nu_delta_init = 500,
                              step_susie = 20,
                              step_R = 30,
                              step_nu = 10,
                              max_iter = 30, 
                              tol = 1e-4)

b_res <- love_res_WS$b_res
R_q   <- love_res_WS$R_q

susie_plot_alpha(
  alpha = b_res$alpha,
  R = cov2cor(R_q),
  b = beta_true,
  main = paste0("SuSiE-LovE with WS shrinkage")
)

```
Let's try again with fixed and small $\nu_0$: It indeed leads to conservative behavior. But how to produce this small $\nu_0$???

```{r}
nu_q = 10
nu_0 = 1
lambda = 1 / N0
susie_love_small_nu <- susie_love_fix_2params(v, R_out, diag_R,
                                   lambda, nu_q, nu_0,
                                   N, L = 5,
                                   k,
                                   z_init = NULL,
                                   step_susie = 20,
                                   step_R = 30,
                                   step_nu = 30,
                                   max_iter = 100,
                                   max_iter_init = 100,
                                   tol = 1e-4)

b_res <- susie_love_small_nu$b_res
R_q   <- susie_love_small_nu$R_q

susie_plot_alpha(
  alpha = b_res$alpha,
  R = cov2cor(R_q),
  b = beta_true,
  main = paste0("SuSiE-LovE with small nu0 = ", nu_0)
)
```

# 3. Trying to achieve conservative behavior when the mismatch is large

Trying a new algorithmic approach:

1) First train the model with some fix $\lambda$ and $\nu_0$

2) Get $R_q$

3) Treat this as a proxy of the true unknown in-sample LD. 

4) Get $v_0 = v - R_q \overline{b}$ and treat this as the null observation to obtain an estimate for $\lambda$.

5) For each column of $R_q$ corresponds to highest PIP variant in each CS, use algorithm `optimize_params_single_effect` to learn $\nu_0$ and $\lambda$.

6) Repeat this process with new $\nu_0$ and $\lambda$.


```{r}
nu_q = N0
nu_0 = N0
lambda = 1 / (N0 + 1)

susie_love_init <- susie_love_fix_2params(v, R0, diag_R,
                                   lambda, nu_q, nu_0,
                                   N, L = 5,
                                   k,
                                   z_init = NULL,
                                   step_susie = 20,
                                   step_R = 30,
                                   step_nu = 30,
                                   max_iter = 100,
                                   max_iter_init = 100,
                                   tol = 1e-4)

b_res <- susie_love_init$b_res
R_q   <- susie_love_init$R_q

susie_plot_alpha(
  alpha = b_res$alpha,
  R = cov2cor(R_q),
  b = beta_true,
  main = paste0("SuSiE-LovE with fixed nu0 = ", nu_0)
)
```

Null estimate

```{r}
b_bar = colSums(b_res$alpha * b_res$mu)
v0 = v - R_q %*% b_bar

null_res = maximize_log_pdf_null(R0, v0, N)
null_res
```

Signal estimate

```{r}
lambda_est = null_res$lambda

R_bar = (1 - lambda_est) * R0 + lambda_est * diag(1, J)

set_rows = c()
for (l in 1:nrow(b_res$alpha)) {
  indx = which.max(b_res$alpha[l, ])
  print(indx)
  set_rows = c(set_rows, indx)
}

for (row in set_rows){
   signal_res = optimize_nu0_single_effect(R_q[row, ], row, R0, init_nu0 = 50, lambda = lambda_est)
   print(signal_res)
}

```


```{r}
for (row in set_rows){
   signal_res = optimize_nu0_single_effect(R[row, ], row, R0, init_nu0 = 50, lambda = lambda_est)
   print(signal_res)
}

```


```{r}
for (row in set_rows){
   signal_res = optimize_params_single_effect(R_q[row, ], row, R0, init_nu0 = 100, lambda_est)
   print(signal_res)
}
```

```{r}
for (row in set_rows){
   grid_res = grid_search_single_effect(R_q[row, ], row, R0, n_grid = 10)
   print(row)  
   print(grid_res$lambda_opt)
   print(grid_res$nu_0_opt)
}



```

```{r}
prior_mean = N0 / (N0 + 1) * R0 + 1 / (N0 + 1) * diag_R
all_dists = colSums((R_q - prior_mean)^2)

plot(c(1:J), all_dists)  
all_cs = susielove:::get_cs(b_res$alpha, R_q)$cs
for (i in c(1:5)){
  points(all_cs[[i]], all_dists[all_cs[[i]]], col=i+1, pch = 19)
}


```

```{r}
prior_mean = N0 / (N0 + 1) * R0 + 1 / (N0 + 1) * diag_R
all_dists = colSums((R - prior_mean)^2)

plot(c(1:J), all_dists)  
all_cs = susielove:::get_cs(b_res$alpha, R_q)$cs
for (i in c(1:5)){
  points(all_cs[[i]], all_dists[all_cs[[i]]], col=i+1, pch = 19)
}


```

# 4. But is it too conservative in other cases???

Now consider another case where the mismatch is low and the true $L = 2$. Let's see if it can recover both signals (and set $\nu_0$ to be large, $\lambda$ to be small) or it is still conservative. 

```{r}
seed  <- 1
T_split <- 1  ## divergence time very small so the mismatch is small
chrom <- "chr22"
start <- 20e6
end   <- 21e6
N     <- 25000L
N0    <- 1000L
J     <- 500L
h2    <- 0.01

res <- sim_2pop(chrom, start, end,
                T_split = T_split,
                n_gwas = N,
                n_ref = N0,
                J = J,
                h2 = h2,
                seed = seed)

v               <- res$v
R               <- res$R
R0              <- res$R0
beta_true       <- res$beta_true
causal_SNPs_loc <- res$causal_SNPs_loc

## standardized
d = 1 / sqrt(diag(R))
v = d * v
R = cov2cor(R)
R0 = cov2cor(R0)
  
plot(v, pch = 16)
points(causal_SNPs_loc, v[causal_SNPs_loc], pch = 16, col = "red")
```

Here is susie fitted with the out-of-sample LD:

```{r}
fit_R_init <- susie_suff_stat(
  XtX = N * R0,
  Xty = N * v,
  n = N,
  yty = N,
  L = 5,
  residual_variance = 1
)

susie_plot(fit_R_init, y = "PIP", b = beta_true,
           main = "Out-of-sample corr")
```

Even the out-of-sample LD works well. We know that the mismatch is small.

```{r}
nu_q = N0
nu_0 = N0
lambda = 1 / (N0 + 1)
susie_love_init <- susie_love_fix_2params(v, R0, diag_R,
                                   lambda, nu_q, nu_0,
                                   N, L = 5,
                                   k,
                                   z_init = NULL,
                                   step_susie = 20,
                                   step_R = 30,
                                   step_nu = 30,
                                   max_iter = 100,
                                   max_iter_init = 100,
                                   tol = 1e-4)

b_res <- susie_love_init$b_res
R_q   <- susie_love_init$R_q

susie_plot_alpha(
  alpha = b_res$alpha,
  R = cov2cor(R_q),
  b = beta_true,
  main = paste0("SuSiE-LovE with fixed nu0 = ", nu_0)
)
```

Null estimate

```{r}
b_bar = colSums(b_res$alpha * b_res$mu)
v0 = v - R_q %*% b_bar

null_res = maximize_log_pdf_null(R0, v0, N)
null_res
```

Signal estimate

```{r}
set_rows = c()
for (l in 1:nrow(b_res$alpha)) {
  indx = which.max(b_res$alpha[l, ])
  print(indx)
  set_rows = c(set_rows, indx)
}
```


```{r}
for (row in set_rows){
   signal_res = optimize_params_single_effect(R_q[row, ], row, R0, init_nu0 = 20, init_lambda = null_res$lambda)
   print(signal_res)
}
```


# 5. Conclusion now.

1) estimating nu_0 from R_q is not working. (Compared to R)

2) For selling, perhaps I should just set $\delta = .1$ and then tune $\nu_0$ in `susie-XL`. That seems most principle.


