---
title: "nu_0 decreases as mismatch increases"
author: "dodat-stats"
date: "2026-06-04"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 1. Aim

We are going to investigate if the amount of mismatch increases then the the estimated $\nu_0$ decreases? We will increase the divergence time in the coalescent simulation.


Recall 3 shrinkage (choosing $\lambda$) rules:

1. Naive shrinkage: 
$$\Psi_{naive} = \left(1-\dfrac{1}{N_0}) R_0 + \dfrac{1}{N_0} I$$

2. Wen and Stephens (2010) shrinkage: 
$$\Psi_{WS} = \left(1- \lambda)^2 R_0 + (1 - (1-\lambda)^2) I,$$
where $\lambda = \dfrac{\theta}{\theta + 2 N_0}$ and $\theta$ is the Watterson's estimator $\theta = \left(\sum_{i=1}^{2N_0 - 1} \dfrac{1}{i}\right)^{-1} \approx (\log(2N_0) + 0.57)^{-1} $.

3. Ledoit and Wolf (2004) linear shrinkage: $$\Psi_{linear-LW} = \left(1- \lambda) R_0 + \lambda I,$$ where $\lambda \asymp \dfrac{1}{N_0}$. 


```{r}


compute_wen_stephens <- function(R_0, N_ref) {
  # Number of haplotypes (diploid population)
  d <- 2 * N_ref
  
  # Calculate Watterson's estimator constant (Harmonic sum)
  harmonic_sum <- sum(1 / (1:(d - 1)))
  theta        <- 1 / harmonic_sum
  
  # Calculate the exact mutation shrinkage parameter lambda
  lambda <- theta / (theta + 2 * N_ref)
  
  # Apply local uniform shrinkage: Psi = (1-lambda)^2 * R_0 + (1 - (1-lambda)^2) * I
  p <- ncol(R_0)
  R_ws <- (1 - lambda)^2 * R_0 + (1 - (1 - lambda)^2) * diag(p)
  
  return(R_ws)
}


compute_lw_linear_from_R0 <- function(R_0, N_ref) {
  p <- ncol(R_0)
  
  # 1. Sum of the squared off-diagonal elements of R_0
  # (This is the Frobenius norm of the off-diagonals)
  sum_sq_off_diag <- sum(R_0^2) - p
  
  # 2. Estimate the sampling variance (noise) parameter under normality
  # In correlation space, the sum of variances of r_ij is approximated by:
  beta_sq <- (1 / N_ref) * (p + sum_sq_off_diag)
  
  # 3. Calculate the distance to the identity target matrix
  # Since diag(R_0) is all 1s and diag(I) is all 1s, the distance is just 
  # the sum of the squared off-diagonal elements.
  delta_sq <- sum_sq_off_diag
  
  # 4. Compute the optimal shrinkage intensity alpha
  # Guard against division by zero if R_0 is already the identity matrix
  if (delta_sq == 0) {
    alpha <- 1
  } else {
    alpha <- min(1, beta_sq / delta_sq)
  }
  
  # 5. Construct the shrunk matrix: alpha * I + (1 - alpha) * R_0
  R_lw_linear <- (1 - alpha) * R_0 + alpha * diag(p)
  
  # Print the learned alpha for transparency
  cat("Ledoit-Wolf Linear Alpha estimated from R_0:", alpha, "\n")
  
  return(R_lw_linear)
}


```


# 1. Running and comparing 3 methods on many simulation

See the end of the result.

```{r simulate}
library(susielove)
library(susieR)

num_reps = 25
coverages = matrix(0, nrow = num_reps, ncol = 5)
powers = matrix(0, nrow = num_reps, ncol = 5)
no_CSs = matrix(0, nrow = num_reps, ncol = 5)
no_SNPs_CS = matrix(0, nrow = num_reps, ncol = 5)
dist_R = matrix(0, nrow = num_reps, ncol = 4)

est_nu_0 = matrix(0, nrow = num_reps, ncol = 3)



chrom <- "chr22"
start <- 20e6
end   <- 21e6
N     <- 25000L
J     <- 1000L
h2    <- 0.01

for (seed in c(1:num_reps)){
  print(paste0("Seed ", seed))
  
  T_split = seed * 20
  N0 = 200
  # N0 = 100 + (num_reps - seed) * 10
  
  res <- sim_2pop(chrom, start, end,
                T_split = T_split,
                n_gwas = N,
                n_ref = N0,
                J = J,
                h2 = h2,
                seed = seed)

  v               <- res$v
  R               <- res$R
  R0              <- res$R0
  beta_true       <- res$beta_true
  causal_SNPs_loc <- res$causal_SNPs_loc
  
  ## standardized
  d = 1 / sqrt(diag(R))
  v = d * v
  R = cov2cor(R)
  R0 = cov2cor(R0)
  
  ## 1. in-sample LD
  fit <- susie_suff_stat(
    XtX = N * R,
    Xty = N * v,
    n = N,
    yty = N
  )
  
  ## 2. Out-of-sample LD
  fit_R0 <- susie_suff_stat(
    XtX = N * R0,
    Xty = N * v,
    n = N,
    yty = N,
    L = 5
  )
  dist_R[seed, 1] = sum((R0 - R)^2)
  
  ## 3. Naive shrinkage
  lambda <- 1 / (N0 + 1)
  diag_R <- diag(R)
  R_out  <- R0
  k      <- min(N0, J)
  
  love_res <- susie_love_lambda(v, R_out, diag_R,
                                lambda,
                                N, L = 5,
                                k,
                                nu_delta_init = 1000,
                                step_susie = 20,
                                step_R = 30,
                                step_nu = 10,
                                max_iter = 30, 
                                tol = 1e-4)
  est_nu_0[seed, 1] = love_res$nu_delta
  dist_R[seed, 2] = sum((love_res$R_q - R)^2)
  
  ## 4. WS shrinkage
  diag_R <- diag(R)
  R_out  <- compute_wen_stephens(R_0 = R0, N_ref = N0)
  k      <- min(N0, J)
  
  love_res_WS <- susie_love_lambda(v, R_out, diag_R,
                                lambda = 0,
                                N, L = 5,
                                k,
                                nu_delta_init = 1000,
                                step_susie = 20,
                                step_R = 30,
                                step_nu = 10,
                                max_iter = 30, 
                                tol = 1e-4)
  
  est_nu_0[seed, 2] = love_res_WS$nu_delta
  dist_R[seed, 3] = sum((love_res_WS$R_q - R)^2)
  ## 5. LW shrinkage
  diag_R <- diag(R)
  R_out  <- compute_lw_linear_from_R0(R0, N0)
  k      <- min(N0, J)
  
  love_res_LW_linear <- susie_love_lambda(v, R_out, diag_R,
                                lambda = 0,
                                N, L = 5,
                                k,
                                nu_delta_init = 1000,
                                step_susie = 20,
                                step_R = 30,
                                step_nu = 10,
                                max_iter = 30, 
                                tol = 1e-4)
  
  est_nu_0[seed, 3] = love_res_LW_linear$nu_delta
  dist_R[seed, 4] = sum((love_res_LW_linear$R_q - R)^2)
  
  fitted_rss = list(fit, 
                    fit_R0, 
                    love_res, 
                    love_res_WS, 
                    love_res_LW_linear)
  
  for (method in 1:5){
    if (method < 3) {
      summary_all_CS = summary(fitted_rss[[method]])$cs
      all_CS <- lapply(summary_all_CS$variable, function(x) {
        as.numeric(unlist(strsplit(x, ",")))})
    } else {
      all_CS = get_cs(fitted_rss[[method]]$b_res$alpha, fitted_rss[[method]]$R_q)$cs
    }
    
    L_infer = length(all_CS)
    if (is.null(all_CS) || length(all_CS) == 0) {
      coverages[seed, method] = 0
      powers[seed, method] = 0
      no_CSs[seed, method] = 0
      no_SNPs_CS[seed, method] = 0
    } else{
      num_contains = 0
      all_selected_SNPs = c()
      for (ell in 1:L_infer){
        this_CS = all_CS[[ell]]
        all_selected_SNPs = c(all_selected_SNPs, this_CS)
        num_contains = num_contains + (length(intersect(this_CS, causal_SNPs_loc)) > 0)
      }
      ## coverage = proportion of CS that contains a true casual SNP
      coverages[seed, method] = num_contains / L_infer
      ## power = proportion of casual SNP that is contained in a CS
      selected = length(intersect(all_selected_SNPs, causal_SNPs_loc)) 
      powers[seed, method] = selected / length(causal_SNPs_loc)
      
      ## number of CSs
      no_CSs[seed, method] = L_infer
      
      ## number of SNPs per CSs
      no_SNPs_CS[seed, method] = length(all_selected_SNPs) / L_infer
    }
  }
}


```

# 2. Visualization results

## 2.1. Estimated nu_0

```{r results 1, fig.width=8, fig.height=6}
library(ggplot2)
df <- data.frame(
  x = c(1:seed) * 10,
  y = est_nu_0[, 1]
)

ggplot(df, aes(x = x, y = y)) +
  geom_point() +                          
  geom_smooth(method = "lm", se = TRUE) + 
  labs(
    x     = "Divergence time",
    y     = "Estimated nu_0",
    title = "naive shrinkage prior mean"
  )
```


```{r results 2, fig.width=8, fig.height=6}
df <- data.frame(
  x = c(1:seed) * 10,
  y = est_nu_0[, 2]
)

ggplot(df, aes(x = x, y = y)) +
  geom_point() +                          
  geom_smooth(method = "lm", se = TRUE) + 
  labs(
    x     = "Divergence time",
    y     = "Estimated nu_0",
    title = "WS shrinkage prior mean"
  )
```

```{r results 3, fig.width=8, fig.height=6}
df <- data.frame(
  x = c(1:seed) * 10,
  y = est_nu_0[, 3]
)

ggplot(df, aes(x = x, y = y)) +
  geom_point() +                          
  geom_smooth(method = "lm", se = TRUE) + 
  labs(
    x     = "Divergence time",
    y     = "Estimated nu_0",
    title = "LW shrinkage prior mean"
  )
```


## 2.2. Number of CSs
```{r num_CS, fig.width=8, fig.height=6}

library(ggplot2)
library(tidyr)
library(dplyr)

colnames(coverages) <- c("In-sample", "Out-of-sample", "Naive shrinkage", "WS", "LW")


# Build a data frame and pivot to long format
df <- data.frame(
  x = c(1:seed) * 10,
  coverages
) %>%
  pivot_longer(
    cols = -x,            # everything except x
    names_to  = "group",  # column name becomes the grouping variable
    values_to = "y"       # the cell values
  )

ggplot(df, aes(x = x, y = y, color = group)) +
  geom_line() +
  geom_point() +
  labs(
    x     = "Divergence time",
    y     = "Coverage",
    title = "Coverage as divergence time varies",
    color = "Legend Title"
  )
```

Because I have not ran replications, it is a bit hard to see. But it looks like WS wins overall (by a small margin).

```{r compare mean coverage}
colSums(coverages) / 25
```

Finally we can visualize the distance from out-of-sample $R_0$ and other estimated $R_q$ to the in-sample $R$. $R_q$ estimated from all methods are better than $R_0$.

```{r dist-R}
colnames(dist_R) <- c("Out-of-sample", "Naive shrinkage", "WS", "LW")


# Build a data frame and pivot to long format
df <- data.frame(
  x = c(1:seed) * 10,
  dist_R
) %>%
  pivot_longer(
    cols = -x,            # everything except x
    names_to  = "group",  # column name becomes the grouping variable
    values_to = "y"       # the cell values
  )

ggplot(df, aes(x = x, y = y, color = group)) +
  geom_line() +
  geom_point() +
  labs(
    x     = "Divergence time",
    y     = "Distance from estimated R to in-sample R",
    title = "Estimation error of R as divergence time varies",
    color = "Legend Title"
  )
```

## 3. Conclusion

We learned that as the divergence between in-sample $R$ and out-of-sample $R_0$ increases, the estimated $\nu_0$ decreases, and the estimated $R_q$ is often a little closer to the in-sample $R$ (compared to the distance between $R_0$ and $R$). (Note that the rows corresponds to true causal signals often have biggest improvements, as observed from experiments that Peter and I conducted earlier.)



