---
title: "Choosing rank k: eigenvalue elbow, KL cliff, and fit error"
author: "Dat Do"
date: "2026-06-16"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# 1. Aim

The rank $k$ of the low-rank part of $R_q = FF^\top + D$ controls a fundamental
tension in `susie_XL`:

- **Too small**: the variational family cannot approximate the prior mean
  $\bar{R}$, so $\mathrm{KL}(q \| \text{prior})$ is huge. Because this KL grows
  with $\nu_0$, it penalises large $\nu_0$ and forces the learned degree-of-freedom
  downward — which is actually the *correct* behaviour under LD mismatch.
- **Too large**: $R_q$ can exactly match $\bar{R} \approx R_0$, the KL collapses
  to zero, the ELBO is dominated by the SuSiE fit term, and $\nu_0$ inflates
  without correction — producing false positives.

We investigate four candidate rules for choosing $k$ automatically:

1. **Exp A — Eigenvalue elbow**: scree plot of $R_0$; pick $k$ where eigenvalues
   become negligible (95 % / 99 % cumulative variance thresholds).
2. **Exp B — KL cliff**: fix $\nu = N_0$ and sweep $k$; find the rank where
   $\mathrm{KL}(q \| \text{prior})$ drops sharply.
3. **Exp C — Fit error**: for each $k$, run SuSiE with $R_q$ as the LD matrix,
   then compute the squared residual $N \|v - R_q \hat{b}\|^2 / J$.
   This measures how well the fine-mapping fits the observed signal under each
   rank constraint.
4. **Exp D — Chosen L**: for each $k$, count the number of credible sets SuSiE
   returns when fit with $R_q$. Tracks how rank constrains SuSiE's ability to
   resolve distinct signals.

Both experiments are run for:

- **Low-mismatch** (`sim_2pop`, T_split = 50, $J = 500$, $N_0 = 500$)
- **High-mismatch** (`sim_asn_eur`, CHB GWAS + EUR reference, $J = 500$, $N_0 = 500$)

# 2. Helper functions

```{r helpers}
library(susielove)
library(susieR)

# KL(IW(nu*R_q, nu+J+1) || IW(nu*R_bar, nu+J+1)) at nu_q = nu_0 = nu
# = (nu+J+1)/2 * [log|R_q| - log|R_bar| + tr(R_bar R_q^{-1}) - J]
compute_KL_IW <- function(Z, k, nu, R_bar, diag_R) {
  J          <- nrow(R_bar)
  s          <- 1 + rowSums(Z^2)
  alpha_diag <- sqrt(diag_R / s)
  F_mat      <- Z * alpha_diag
  D_inv_diag <- s / diag_R

  M            <- diag(k) + t(F_mat) %*% (F_mat * D_inv_diag)
  chol_M       <- chol(M)
  log_det_Rq   <- sum(log(diag_R / s)) + 2 * sum(log(diag(chol_M)))
  log_det_Rbar <- 2 * sum(log(diag(chol(R_bar))))

  M_inv      <- chol2inv(chol_M)
  D_inv_F    <- F_mat * D_inv_diag
  term1      <- sum(diag(R_bar) * D_inv_diag)
  A          <- crossprod(D_inv_F, R_bar %*% D_inv_F)
  tr_Rbar_Rq <- term1 - sum(A * M_inv)

  (nu + J + 1) / 2 * (log_det_Rq - log_det_Rbar + tr_Rbar_Rq - J)
}

# SuSiE-based fit error and chosen L: run susie_suff_stat with R_q as LD,
# return N * ||v - R_q b_hat||^2 / J and number of credible sets
compute_fit_susie <- function(Z, k, diag_R, v, N, L = 5) {
  s          <- 1 + rowSums(Z^2)
  alpha_diag <- sqrt(diag_R / s)
  F_mat      <- Z * alpha_diag
  D_diag     <- alpha_diag^2

  R_q <- tcrossprod(F_mat) + diag(D_diag)

  fit_xl <- suppressWarnings(
    susieR::susie_suff_stat(XtX = N * R_q, Xty = N * v,
                             n = N, yty = N, L = L)
  )
  b_hat   <- colSums(fit_xl$alpha * fit_xl$mu)
  epsilon <- v - R_q %*% b_hat
  n_cs    <- length(fit_xl$sets$cs)

  list(fit_error = N * sum(epsilon^2) / length(v), n_cs = n_cs)
}

# Sweep k: compute KL, SuSiE fit error, and chosen L for each value in k_grid
run_k_sweep <- function(R_bar, nu_fixed, diag_R, v, k_grid, N,
                        max_iter_init = 200, L = 5) {
  J       <- length(v)
  results <- data.frame(k = k_grid, KL = NA_real_, fit_error = NA_real_,
                        n_cs = NA_integer_)
  for (i in seq_along(k_grid)) {
    ki        <- k_grid[i]
    z         <- susielove:::initialize_Z_from_prior_R(
                   R_bar, nu_fixed, nu_fixed, diag_R, ki,
                   max_iter_init = max_iter_init)
    Z         <- matrix(z, nrow = J, ncol = ki)
    susie_res <- compute_fit_susie(Z, ki, diag_R, v, N, L = L)
    results$KL[i]         <- compute_KL_IW(Z, ki, nu_fixed, R_bar, diag_R)
    results$fit_error[i]  <- susie_res$fit_error
    results$n_cs[i]       <- susie_res$n_cs
  }
  results
}

# k at which cumulative % explained variance >= threshold
eigen_k_thresh <- function(eig_vals, threshold) {
  pct_cum <- cumsum(eig_vals) / sum(eig_vals)
  which(pct_cum >= threshold)[1]
}

# k right after the largest single-step drop in log(KL)
kl_cliff_k <- function(sweep_df) {
  log_kl <- log(pmax(sweep_df$KL, 1e-10))
  drops  <- diff(log_kl)
  sweep_df$k[which.min(drops) + 1]
}
```

---

# 3. Low-mismatch scenario

Same-population two-epoch split (`sim_2pop`, T_split = 25, $N_0 = 500$, $J = 500$).

## 3.1 Simulate data

```{r sim-low, fig.width=8, fig.height=5}
seed    <- 1
T_split <- 50
chrom   <- "chr22"
start   <- 20e6
end     <- 21e6
N       <- 25000L
N0      <- 500L
J       <- 500L
h2      <- 0.01

res_low        <- sim_2pop(chrom, start, end,
                            T_split = T_split,
                            n_gwas  = N,
                            n_ref   = N0,
                            J       = J,
                            h2      = h2,
                            seed    = seed)
v_low      <- res_low$v
R_low      <- res_low$R
R0_low     <- res_low$R0
beta_low   <- res_low$beta_true
causal_low <- res_low$causal_SNPs_loc

# diagonal correction (match test-new-optimization-delta-ver)
d0_inv_sqrt <- 1 / sqrt(diag(R0_low))
LD0         <- diag(d0_inv_sqrt) %*% R0_low %*% diag(d0_inv_sqrt)
d_sqrt      <- sqrt(diag(R_low))
R_tilde_low <- diag(d_sqrt) %*% LD0 %*% diag(d_sqrt)

diag_R_low <- diag(R_low)
R_out_low  <- R_tilde_low

# prior mean at nu_0 = N0, delta = 1
R_bar_low <- (1 - 1/N0) * R_out_low + (1/N0) * diag(diag_R_low)
nu_low    <- N0

plot(v_low, pch = 16, cex = 0.4,
     main = "Summary statistics — low mismatch (T_split=25)")
points(causal_low, v_low[causal_low], pch = 16, col = "red", cex = 1.2)
```

## 3.2 Exp A — Eigenvalue elbow of $R_{\text{out}}$

```{r eig-low}
eig_low  <- eigen(R_out_low, symmetric = TRUE, only.values = TRUE)$values
eig_low  <- pmax(eig_low, 0)

k_95_low <- eigen_k_thresh(eig_low, 0.95)
k_99_low <- eigen_k_thresh(eig_low, 0.99)

cat("k at 95% variance:", k_95_low, "\n")
cat("k at 99% variance:", k_99_low, "\n")
```

## 3.3 Exp B — KL cliff and Exp C — SuSiE fit error vs $k$

We fix $\nu = N_0 =$ `r N0` and sweep $k$ up to $\min(N_0, J) - 10 =$ `r min(N0, J) - 10`.
For each $k$ we find the best rank-$k$-plus-diagonal $R_q$ via
`initialize_Z_from_prior_R`, compute the IW KL divergence, and run SuSiE
with that $R_q$ to compute $N\|v - R_q\hat{b}\|^2 / J$.

```{r sweep-low}
k_grid_low <- c(10, 25, 50, 75, 100, 125, 150, 200, 250, 300, 350, 400, 450, 490)

sweep_low <- run_k_sweep(R_bar_low, nu_low, diag_R_low, v_low,
                          k_grid_low, N, max_iter_init = 200)
print(sweep_low)
```

```{r plot-scree-kl-low, fig.width=10, fig.height=4}
par(mfrow = c(1, 2))

plot(seq_along(eig_low), eig_low, pch = 16, cex = 0.35,
     xlab = "Index", ylab = "Eigenvalue",
     main = "Scree plot — low mismatch")
abline(v = k_95_low, col = "red",       lty = 2)
abline(v = k_99_low, col = "darkgreen", lty = 2)
legend("topright",
       legend = c(paste0("95%: k=", k_95_low),
                  paste0("99%: k=", k_99_low)),
       col = c("red", "darkgreen"), lty = 2, cex = 0.8)

plot(sweep_low$k, sweep_low$KL, type = "b", pch = 16,
     xlab = "k", ylab = "KL(q || prior)",
     main = "KL vs k — low mismatch")
abline(v = k_95_low, col = "red",       lty = 2)
abline(v = k_99_low, col = "darkgreen", lty = 2)
legend("topright",
       legend = c(paste0("95% eig: k=", k_95_low),
                  paste0("99% eig: k=", k_99_low)),
       col = c("red", "darkgreen"), lty = 2, cex = 0.8)

par(mfrow = c(1, 1))
```

## 3.4 Exp C+D — Fit error and Chosen L

```{r plot-fit-cs-low, fig.width=10, fig.height=4}
par(mfrow = c(1, 2))

plot(sweep_low$k, sweep_low$fit_error, type = "b", pch = 16,
     xlab = "k", ylab = "N * ||v - R_q * b_hat||^2 / J",
     main = "SuSiE fit error vs k — low mismatch")
abline(v = k_95_low, col = "red",       lty = 2)
abline(v = k_99_low, col = "darkgreen", lty = 2)
legend("topright",
       legend = c(paste0("95% eig: k=", k_95_low),
                  paste0("99% eig: k=", k_99_low)),
       col = c("red", "darkgreen"), lty = 2, cex = 0.8)

plot(sweep_low$k, sweep_low$n_cs, type = "b", pch = 16,
     xlab = "k", ylab = "Number of credible sets",
     main = "Chosen L vs k — low mismatch",
     ylim = c(0, max(sweep_low$n_cs, na.rm = TRUE) + 1))
abline(v = k_95_low, col = "red",       lty = 2)
abline(v = k_99_low, col = "darkgreen", lty = 2)

par(mfrow = c(1, 1))
```

## 3.5 Summary — low mismatch

```{r summary-low}
k_kl_low <- kl_cliff_k(sweep_low)

cat("Eigenvalue 95%:   k =", k_95_low, "\n")
cat("Eigenvalue 99%:   k =", k_99_low, "\n")
cat("KL cliff:         k =", k_kl_low, "\n")
cat("Fit error range: [", round(min(sweep_low$fit_error), 2),
    ",", round(max(sweep_low$fit_error), 2), "]\n")
cat("Fit error % variation:",
    round(100 * diff(range(sweep_low$fit_error)) / mean(sweep_low$fit_error), 1),
    "%\n")
```

---

# 4. High-mismatch scenario

CHB GWAS + EUR reference (`sim_asn_eur`, $N_0 = 500$, $J = 500$). Mirrors
`mismatch_fix_delta_approach`.

## 4.1 Simulate data

```{r sim-asn-eur-fn}
sim_asn_eur <- function(chrom, start, end, n_gwas, n_ref, J, h2, seed) {
  stdpopsim <- reticulate::import("stdpopsim")
  msprime   <- reticulate::import("msprime")
  set.seed(seed)

  species    <- stdpopsim$get_species("HomSap")
  contig     <- species$get_contig(chrom,
                                    left  = as.integer(start),
                                    right = as.integer(end))
  demo_model <- species$get_demographic_model("OutOfAfrica_3G09")
  demography <- demo_model$model

  ts <- msprime$sim_ancestry(
    samples = reticulate::dict(CHB = as.integer(n_gwas), CEU = as.integer(n_ref)),
    demography = demography,
    recombination_rate = contig$recombination_map,
    random_seed = as.integer(seed)
  )
  ts <- msprime$sim_mutations(ts, rate = 1.29e-8, random_seed = as.integer(seed))

  chb_id <- NULL; ceu_id <- NULL
  for (p in reticulate::iterate(ts$populations())) {
    if (!is.null(p$metadata) && p$metadata$name == "CHB") chb_id <- p$id
    if (!is.null(p$metadata) && p$metadata$name == "CEU") ceu_id <- p$id
  }
  G_all        <- ts$genotype_matrix()
  samples_gwas <- ts$samples(population = as.integer(chb_id))
  samples_ref  <- ts$samples(population = as.integer(ceu_id))

  to_diploid <- function(hap_idx) {
    idx   <- as.integer(hap_idx) + 1
    g_hap <- G_all[, idx, drop = FALSE]
    t(g_hap[, seq(1, ncol(g_hap), 2)] + g_hap[, seq(2, ncol(g_hap), 2)])
  }
  G_gwas <- to_diploid(samples_gwas)
  G_ref  <- to_diploid(samples_ref)

  maf_gwas    <- colMeans(G_gwas) / 2
  maf_ref     <- colMeans(G_ref)  / 2
  common_mask <- (maf_gwas > 0.05) & (maf_gwas < 0.95) &
                 (maf_ref  > 0.05) & (maf_ref  < 0.95)
  n_common    <- sum(common_mask)
  if (n_common < J)
    stop(paste0("Only ", n_common, " common variants, need J=", J))
  G_gwas <- G_gwas[, common_mask, drop = FALSE]
  G_ref  <- G_ref[,  common_mask, drop = FALSE]

  J_all     <- min(ncol(G_gwas), J)
  max_start <- J_all - J
  start_idx <- if (max_start > 0) sample(0:max_start, 1) else 0
  idx       <- (start_idx + 1):(start_idx + J)
  G  <- G_gwas[, idx, drop = FALSE]
  G0 <- G_ref[,  idx, drop = FALSE]

  X           <- sweep(G,  2, colMeans(G),  "-")
  G0_centered <- sweep(G0, 2, colMeans(G0), "-")
  R0 <- stats::cov(G0_centered); R0[is.na(R0)] <- 0
  R  <- stats::cov(X);           R[is.na(R)]   <- 0

  first_causal    <- sample(1:J, 1)
  second_causal   <- which.min(R[first_causal, ]^2)
  causal_SNPs_loc <- c(first_causal, second_causal)
  beta_true       <- rep(0, J)
  beta_true[causal_SNPs_loc] <- stats::runif(2) + 0.01
  var_expl  <- sum((X %*% beta_true)^2 / nrow(X))
  beta_true <- beta_true * sqrt(h2 / var_expl)

  y     <- (X %*% beta_true) + sqrt(1 - h2) * stats::rnorm(nrow(X))
  y_std <- sqrt(sum((y - mean(y))^2) / nrow(X))
  y     <- (y - mean(y)) / y_std
  v     <- as.vector(t(X) %*% y / nrow(X))

  list(v = v, R = R, R0 = R0, beta_true = beta_true,
       causal_SNPs_loc = causal_SNPs_loc)
}
```

```{r sim-high, fig.width=8, fig.height=5}
N_hi  <- 25000L
N0_hi <- 500L
J_hi  <- 500L

res_high    <- sim_asn_eur("chr22", 20e6, 21e6,
                            n_gwas = N_hi, n_ref = N0_hi,
                            J = J_hi, h2 = 0.01, seed = 2)
v_high      <- res_high$v
R_high      <- res_high$R
R0_high     <- res_high$R0
beta_high   <- res_high$beta_true
causal_high <- res_high$causal_SNPs_loc

# standardise to correlation scale (match mismatch_fix_delta_approach)
d_hi    <- 1 / sqrt(diag(R_high))
v_high  <- d_hi * v_high
R_high  <- cov2cor(R_high)
R0_high <- cov2cor(R0_high)

diag_R_high <- rep(1, J_hi)
R_out_high  <- R0_high
R_bar_high  <- (1 - 1/N0_hi) * R0_high + (1/N0_hi) * diag(J_hi)
nu_high     <- N0_hi

plot(v_high, pch = 16, cex = 0.4,
     main = "Summary statistics — high mismatch (CHB/EUR)")
points(causal_high, v_high[causal_high], pch = 16, col = "red", cex = 1.2)
```

## 4.2 Exp A — Eigenvalue elbow of $R_0$

```{r eig-high}
eig_high  <- eigen(R0_high, symmetric = TRUE, only.values = TRUE)$values
eig_high  <- pmax(eig_high, 0)

k_95_high <- eigen_k_thresh(eig_high, 0.95)
k_99_high <- eigen_k_thresh(eig_high, 0.99)

cat("k at 95% variance:", k_95_high, "\n")
cat("k at 99% variance:", k_99_high, "\n")
```

## 4.3 Exp B — KL cliff and Exp C — SuSiE fit error vs $k$

```{r sweep-high}
k_grid_high <- c(10, 25, 50, 75, 100, 125, 150, 200, 250, 300, 350, 400, 450, 490)

sweep_high <- run_k_sweep(R_bar_high, nu_high, diag_R_high, v_high,
                           k_grid_high, N_hi, max_iter_init = 200)
print(sweep_high)
```

```{r plot-scree-kl-high, fig.width=10, fig.height=4}
par(mfrow = c(1, 2))

plot(seq_along(eig_high), eig_high, pch = 16, cex = 0.35,
     xlab = "Index", ylab = "Eigenvalue",
     main = "Scree plot — high mismatch")
abline(v = k_95_high, col = "red",       lty = 2)
abline(v = k_99_high, col = "darkgreen", lty = 2)
legend("topright",
       legend = c(paste0("95%: k=", k_95_high),
                  paste0("99%: k=", k_99_high)),
       col = c("red", "darkgreen"), lty = 2, cex = 0.8)

plot(sweep_high$k, sweep_high$KL, type = "b", pch = 16,
     xlab = "k", ylab = "KL(q || prior)",
     main = "KL vs k — high mismatch")
abline(v = k_95_high, col = "red",       lty = 2)
abline(v = k_99_high, col = "darkgreen", lty = 2)
legend("topright",
       legend = c(paste0("95% eig: k=", k_95_high),
                  paste0("99% eig: k=", k_99_high)),
       col = c("red", "darkgreen"), lty = 2, cex = 0.8)

par(mfrow = c(1, 1))
```

## 4.4 Exp C+D — Fit error and Chosen L

```{r plot-fit-cs-high, fig.width=10, fig.height=4}
par(mfrow = c(1, 2))

plot(sweep_high$k, sweep_high$fit_error, type = "b", pch = 16,
     xlab = "k", ylab = "N * ||v - R_q * b_hat||^2 / J",
     main = "SuSiE fit error vs k — high mismatch")
abline(v = k_95_high, col = "red",       lty = 2)
abline(v = k_99_high, col = "darkgreen", lty = 2)
legend("topleft",
       legend = c(paste0("95% eig: k=", k_95_high),
                  paste0("99% eig: k=", k_99_high)),
       col = c("red", "darkgreen"), lty = 2, cex = 0.8)

plot(sweep_high$k, sweep_high$n_cs, type = "b", pch = 16,
     xlab = "k", ylab = "Number of credible sets",
     main = "Chosen L vs k — high mismatch",
     ylim = c(0, max(sweep_high$n_cs, na.rm = TRUE) + 1))
abline(v = k_95_high, col = "red",       lty = 2)
abline(v = k_99_high, col = "darkgreen", lty = 2)

par(mfrow = c(1, 1))
```

## 4.5 Summary — high mismatch

```{r summary-high}
k_kl_high <- kl_cliff_k(sweep_high)

cat("Eigenvalue 95%:   k =", k_95_high, "\n")
cat("Eigenvalue 99%:   k =", k_99_high, "\n")
cat("KL cliff:         k =", k_kl_high, "\n")
cat("Fit error range: [", round(min(sweep_high$fit_error), 2),
    ",", round(max(sweep_high$fit_error), 2), "]\n")
cat("Fit error % variation:",
    round(100 * diff(range(sweep_high$fit_error)) / mean(sweep_high$fit_error), 1),
    "%\n")
```

---

# 5. Comparison across scenarios

```{r comparison-table, results='asis'}
tab <- data.frame(
  Criterion     = c("Eigenvalue 95%", "Eigenvalue 99%", "KL cliff"),
  Low_mismatch  = c(k_95_low,  k_99_low,  k_kl_low),
  High_mismatch = c(k_95_high, k_99_high, k_kl_high)
)
knitr::kable(tab,
             col.names = c("Criterion", "Low mismatch k", "High mismatch k"),
             align     = "lcc")
```

```{r comparison-plot, fig.width=10, fig.height=5}
par(mfrow = c(1, 2))

kl_rel_low  <- sweep_low$KL  / sweep_low$KL[1]
kl_rel_high <- sweep_high$KL / sweep_high$KL[1]

plot(sweep_low$k, kl_rel_low, type = "b", pch = 16,
     xlab = "k", ylab = "KL / KL(k=10)",
     main = "Relative KL — low mismatch", ylim = c(0, 1))
abline(v = k_95_low, col = "red",       lty = 2)
abline(v = k_99_low, col = "darkgreen", lty = 2)
abline(v = k_kl_low, col = "purple",    lty = 1, lwd = 2)
legend("topright",
       legend = c(paste0("95% eig: k=", k_95_low),
                  paste0("99% eig: k=", k_99_low),
                  paste0("KL cliff: k=", k_kl_low)),
       col = c("red", "darkgreen", "purple"),
       lty = c(2, 2, 1), cex = 0.8)

plot(sweep_high$k, kl_rel_high, type = "b", pch = 16,
     xlab = "k", ylab = "KL / KL(k=10)",
     main = "Relative KL — high mismatch", ylim = c(0, 1))
abline(v = k_95_high, col = "red",       lty = 2)
abline(v = k_99_high, col = "darkgreen", lty = 2)
abline(v = k_kl_high, col = "purple",    lty = 1, lwd = 2)
legend("topright",
       legend = c(paste0("95% eig: k=", k_95_high),
                  paste0("99% eig: k=", k_99_high),
                  paste0("KL cliff: k=", k_kl_high)),
       col = c("red", "darkgreen", "purple"),
       lty = c(2, 2, 1), cex = 0.8)

par(mfrow = c(1, 1))
```

**Key observations:**

1. **KL needs far more rank than eigenvalues suggest**: the KL cliff occurs at a
   much larger $k$ than either the 95 % or 99 % eigenvalue threshold. This means
   the Frobenius-norm approximation quality of $R_q$ (captured by the eigenvalue
   elbow) is not the right lens — the KL divergence between IW distributions is
   governed by the log-determinant and trace terms, which remain large even when
   the entrywise approximation error is already small.

2. **Euclidean fit error**:  $N\|v - R_q\hat{b}\|^2/J$ as a function of $k$
   reveals how rank constrains SuSiE's ability to explain the observed
   signal. Whether it is monotone, U-shaped, or flat determines whether it can
   complement the KL criterion as a practical selection rule.

3. **Practical implication**: the eigenvalue 95–99 % rule gives a $k$ that is
   safe from the KL cliff (the KL is still large there, so $\nu_0$ is
   penalised appropriately). Using $k$ beyond the 99 % threshold risks entering
   the KL collapse regime and losing mismatch correction.
