---
title: "Silly R0 leads to very small nu0 and conservative estimate?"
author: "dodat-stats"
date: "2026-06-04"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# 1. Aim

We investigate to see if using a **very mismatched** R0 leads to small $\nu_0$.

SPOIL: It actually does not lead to a small $\nu_0$. Ironically, the regularization strength $\lambda$ matters so much more than the discrepancy between $v$ and out-of-sample $R$ in deciding the optimal $\nu_0$... We have had automatic way to decide $\lambda$, but it only depends on $N_0$. Therefore in this case, where $R0$ is way different than $R$, but because $N_0$ is the same, $\lambda$ is the same and leads to quite large $\nu_0$, which leads to more borrowing of information than we like, and leads to more false positive. 

From a bright side, the current implementation is good when $R_0$ is not too mismatched though (mismatch but not super far away from $R$ (like divergence from a few hundred years apart), like in the previous workflow file). 

Anyway, let's see it in action. We generate R from Asian population and R0 from European population.

```{r, include=FALSE}
library(susielove)
library(susieR)

compute_wen_stephens <- function(R_0, N_ref) {
  # Number of haplotypes (diploid population)
  d <- 2 * N_ref
  
  # Calculate Watterson's estimator constant (Harmonic sum)
  harmonic_sum <- sum(1 / (1:(d - 1)))
  theta        <- 1 / harmonic_sum
  
  # Calculate the exact mutation shrinkage parameter lambda
  lambda <- theta / (theta + 2 * N_ref)
  
  # Apply local uniform shrinkage: Psi = (1-lambda)^2 * R_0 + (1 - (1-lambda)^2) * I
  p <- ncol(R_0)
  R_ws <- (1 - lambda)^2 * R_0 + (1 - (1 - lambda)^2) * diag(p)
  
  return(R_ws)
}


compute_lw_linear_from_R0 <- function(R_0, N_ref) {
  p <- ncol(R_0)
  
  # 1. Sum of the squared off-diagonal elements of R_0
  # (This is the Frobenius norm of the off-diagonals)
  sum_sq_off_diag <- sum(R_0^2) - p
  
  # 2. Estimate the sampling variance (noise) parameter under normality
  # In correlation space, the sum of variances of r_ij is approximated by:
  beta_sq <- (1 / N_ref) * (p + sum_sq_off_diag)
  
  # 3. Calculate the distance to the identity target matrix
  # Since diag(R_0) is all 1s and diag(I) is all 1s, the distance is just 
  # the sum of the squared off-diagonal elements.
  delta_sq <- sum_sq_off_diag
  
  # 4. Compute the optimal shrinkage intensity alpha
  # Guard against division by zero if R_0 is already the identity matrix
  if (delta_sq == 0) {
    alpha <- 1
  } else {
    alpha <- min(1, beta_sq / delta_sq)
  }
  
  # 5. Construct the shrunk matrix: alpha * I + (1 - alpha) * R_0
  R_lw_linear <- (1 - alpha) * R_0 + alpha * diag(p)
  
  # Print the learned alpha for transparency
  cat("Ledoit-Wolf Linear Alpha estimated from R_0:", alpha, "\n")
  
  return(R_lw_linear)
}


```


# 2. An example

## 2.1. Generate silly R0

```{r}
sim_asn_eur <- function(chrom, start, end, n_gwas, n_ref, J, h2, seed) {

  stdpopsim <- reticulate::import("stdpopsim")
  msprime   <- reticulate::import("msprime")

  set.seed(seed)

  # 1. Get Realistic Genetic Map from stdpopsim
  species <- stdpopsim$get_species("HomSap")
  contig  <- species$get_contig(chrom, left = as.integer(start), right = as.integer(end))

  # 2. Get OutOfAfrica_3G09 Demography (YRI, CEU, CHB)
  demo_model <- species$get_demographic_model("OutOfAfrica_3G09")
  demography <- demo_model$model

  # 3. Run Simulation
  ts <- msprime$sim_ancestry(
    samples = reticulate::dict(CHB = as.integer(n_gwas), CEU = as.integer(n_ref)),
    demography = demography,
    recombination_rate = contig$recombination_map,
    random_seed = as.integer(seed)
  )

  ts <- msprime$sim_mutations(ts, rate = 1.29e-8, random_seed = as.integer(seed))

  # 4. Extract Genotypes
  chb_id <- NULL
  ceu_id <- NULL
  for (p in reticulate::iterate(ts$populations())) {
    if (!is.null(p$metadata) && p$metadata$name == "CHB") chb_id <- p$id
    if (!is.null(p$metadata) && p$metadata$name == "CEU") ceu_id <- p$id
  }

  samples_gwas <- ts$samples(population = as.integer(chb_id))
  samples_ref  <- ts$samples(population = as.integer(ceu_id))

  G_all <- ts$genotype_matrix()

  to_diploid <- function(hap_idx) {
    idx <- as.integer(hap_idx) + 1
    g_hap <- G_all[, idx, drop = FALSE]
    g_dip <- g_hap[, seq(1, ncol(g_hap), by = 2)] + g_hap[, seq(2, ncol(g_hap), by = 2)]
    return(t(g_dip))
  }

  G_gwas <- to_diploid(samples_gwas)
  G_ref  <- to_diploid(samples_ref)

  # 5. Filter for Common Variants (MAF > 5%)
  maf_gwas <- colMeans(G_gwas) / 2
  maf_ref  <- colMeans(G_ref) / 2

  common_mask <- (maf_gwas > 0.05) & (maf_gwas < 0.95) & (maf_ref > 0.05) & (maf_ref < 0.95)
  n_common <- sum(common_mask)
  if (n_common < J) stop(paste0("Only ", n_common, " common variants found but J = ", J, ". Try a larger region."))
  G_gwas <- G_gwas[, common_mask, drop = FALSE]
  G_ref  <- G_ref[, common_mask, drop = FALSE]

  J_all <- min(ncol(G_gwas), J)
  max_start <- J_all - J
  start_idx <- if (max_start > 0) sample(0:max_start, 1) else 0
  cols_to_keep <- (start_idx + 1):(start_idx + J)
  G  <- G_gwas[, cols_to_keep, drop = FALSE]
  G0 <- G_ref[, cols_to_keep, drop = FALSE]

  # 6. Generate summary stats
  X <- sweep(G, 2, colMeans(G), "-")
  N <- nrow(X)

  G0_centered <- sweep(G0, 2, colMeans(G0), "-")

  R0 <- stats::cov(G0_centered)
  R0[is.na(R0)] <- 0

  R <- stats::cov(X)
  R[is.na(R)] <- 0

  num_causal <- 2
  first_causal_SNP <- sample(1:J, 1)
  second_causal_SNP <- which.min(R[first_causal_SNP, ]^2)
  causal_SNPs_loc <- c(first_causal_SNP, second_causal_SNP)

  beta_true <- rep(0, J)
  beta_true[causal_SNPs_loc] <- stats::runif(num_causal) + 0.01

  var_expl <- sum((X %*% beta_true)^2 / N)
  scale <- h2 / var_expl
  beta_true <- beta_true * sqrt(scale)

  y <- (X %*% beta_true) + sqrt(1 - h2) * stats::rnorm(N)
  y_std <- sqrt(sum((y - mean(y))^2) / N)
  y <- (y - mean(y)) / y_std

  v <- (t(X) %*% y) / N

  return(list(
    v = as.vector(v),
    R = R,
    R0 = R0,
    beta_true = beta_true,
    causal_SNPs_loc = causal_SNPs_loc
  ))
}

```

```{r}
seed = 2
chrom <- "chr22"
start <- 20e6
end   <- 21e6
N     <- 25000L
N0 = 500L
J     <- 500L
h2    <- 0.01

res <- sim_asn_eur(chrom, start, end,
                n_gwas = N,
                n_ref = N0,
                J = J,
                h2 = h2,
                seed = seed)

v               <- res$v
R               <- res$R
R0              <- res$R0
beta_true       <- res$beta_true
causal_SNPs_loc <- res$causal_SNPs_loc

## standardized
d = 1 / sqrt(diag(R))
v = d * v
R = cov2cor(R)
R0 = cov2cor(R0)

plot(v, pch = 16)
points(causal_SNPs_loc, v[causal_SNPs_loc], pch = 16, col = "red")
```



## 2.2 Visualize two correlation matrices

Sorry the color scheme is a bit hard to see but if we squeeze our eyes a bit then we can see that the two matrices are totally different. 

```{r, fig.width=8, fig.height=6}
par(mfrow = c(1, 2))

image(R0[1:100, 1:100], main = "Out-of-sample (EUR)", col = heat.colors(15))
image(R[1:100, 1:100], main = "In-sample corr (CHB)", col = heat.colors(15))
par(mfrow = c(1, 1))

```

## 2.3 Fit susie with in-sample R

```{r, fig.width=8, fig.height=6}
fit_R <- susie_suff_stat(
  XtX = N * R,
  Xty = N * v,
  n = N,
  yty = N,
  L = 5
)
susie_plot(fit_R, y = "PIP", b = beta_true, main = "In-sample cov mat")
```

## 2.4 Fit susie with silly R

```{r, fig.width=8, fig.height=6}
fit_R_silly <- susie_suff_stat(
  XtX = N * R0,
  Xty = N * v,
  n = N,
  yty = N,
  L = 5,
  residual_variance = 1
)
susie_plot(fit_R_silly, y = "PIP", b = beta_true, main = "Silly corr mat")
```
Slightly regularized

```{r, fig.width=8, fig.height=6}
R_tilde = (1 - 1 / N0) * R0 + 1 / N0 * diag(J)
fit_R0 <- susie_suff_stat(
  XtX = N * R0,
  Xty = N * v,
  n = N,
  yty = N,
  L = 5
)
susie_plot(fit_R_silly, y = "PIP", b = beta_true, main = "Silly corr mat")

print(summary(fit_R0)$cs)
```

## 2.5 Fit susielove with silly R

```{r}
R_out  <- compute_wen_stephens(R_0 = R_tilde, N_ref = N0)
# R_out = diag(J)
# R_out = R_tilde
k      <- min(N0, J)
diag_R = diag(R)

love_res_WS <- susie_love_lambda(v, R_out, diag_R,
                              lambda = 0,
                              N, L = 5,
                              k,
                              nu_delta_init = 500,
                              step_susie = 20,
                              step_R = 30,
                              step_nu = 10,
                              max_iter = 30, 
                              tol = 1e-4)

b_res <- love_res_WS$b_res
R_q   <- love_res_WS$R_q

susie_plot_alpha(
  alpha = b_res$alpha,
  R = cov2cor(R_q),
  b = beta_true,
  main = paste0("SuSiE-LovE with WS shrinkage")
)

```
Let's try again with fixed and small $\nu_0$: It indeed leads to conservative behavior. But how to produce this small $\nu_0$???

```{r}
nu_q = 10
nu_0 = 1
lambda = 1 / N0
susie_love_small_nu <- susie_love_fix_2params(v, R_out, diag_R,
                                   lambda, nu_q, nu_0,
                                   N, L = 5,
                                   k,
                                   z_init = NULL,
                                   step_susie = 20,
                                   step_R = 30,
                                   step_nu = 30,
                                   max_iter = 100,
                                   max_iter_init = 100,
                                   tol = 1e-4)

b_res <- susie_love_small_nu$b_res
R_q   <- susie_love_small_nu$R_q

susie_plot_alpha(
  alpha = b_res$alpha,
  R = cov2cor(R_q),
  b = beta_true,
  main = paste0("SuSiE-LovE with small nu0 = ", nu_0)
)
```

## 3. Trying to achieve conservative behavior when the mismatch is large

As discussed, the current way to choose $\lambda$ only depends on $N_0$, but not on how "mismatched" $R_0$ and $v$ is. 

Going back to our desire to learn $\lambda$ as a device to measure the mismatch between $v$ and $R_0$. We have learned the following:

(A) Learning $\lambda$ under the null model is good.

(B) But under the VI of susie-love, the ELBO prefers large $\lambda$ (and also large $\nu_0$), leading to FDR.

(C) But intuitively, large $\lambda$ means large mismatch, which should lead to small $\nu_0$ instead.

(D) One way to rescue this is to fix $\delta$: Indeed, in this case, $\lambda = \dfrac{\delta}{\delta + \nu_0}$, which is inversely proportional to $\nu_0$ and it makes more sense.

Therefore, we can consider the following procedure of estimating $\delta$ and $\nu_0$ from data:

1) First fit the susieR with the out-of-sample LD $R_0$ and chosen $L$ (say $L = 5$) to get the posterior mean $\bar{b}$, then get the "estimated null" $v_0 = \sqrt{N}(v - R_0 \bar{b})$

2) Estimate $\lambda$ from the MLE of the null model $\sqrt{N} v_0 \sim (0, (1 - \lambda) R_0 + \lambda I)$ (or the moment method; see the file `learn-nu0-lambda.tex` on Overleaf).

3) Set $\nu_0 = \dfrac{\delta (1 - \lambda)}{\lambda}$.

This is still estimating the prior from data (so it still has "empirical Bayes" spirit). Let's see how it works

### 3.1. Trying with severely mismatched data

```{r}
moment_estimate_lambda <- function(v0, R0, N){
  eig = eigen(R0)
  U = eig$vectors
  s = eig$values
  w = sqrt(N) * t(U) %*% v0
  lin_mod = lm(w^2 ~ s)
  slope = lin_mod$coefficients[[2]]  # tau^2 * (1 - lambda)
  intercept = lin_mod$coefficients[[1]]  # tau^2 * lambda
  odds = intercept / slope
  lambda_est = odds / (1 + odds)
  
  lambda_est
}
```


```{r}
maximize_log_pdf <- function(R0, v, N) {
  J <- length(v)
  
  R0 <- (R0 + t(R0)) / 2
  eig <- eigen(R0, symmetric = TRUE)
  
  # Clamp any floating-point negative eigenvalues to exactly 0
  d <- pmax(eig$values, 0) 
  v_tilde <- as.numeric(crossprod(eig$vectors, v))
  
  # Optimize over log_nu0 to keep parameter scales identical
  neg_log_pdf <- function(params) {
    lambda <- params[1]
    nu0 <- exp(params[2]) # Decode log(nu0)
    
    d_lambda <- (1 - lambda) * d + lambda
    
    log_det <- sum(log(d_lambda))
    qf <- N * sum((v_tilde^2) / d_lambda)
    
    inner_log <- 1 + (1 / nu0) * qf
    
    term1 <- lgamma((nu0 + J + 2) / 2)
    term2 <- lgamma((nu0 + 2) / 2)
    term3 <- (J / 2) * log(nu0 * pi)
    term4 <- 0.5 * log_det
    term5 <- ((nu0 + J + 2) / 2) * log(inner_log)
    
    log_lik <- term1 - term2 - term3 - term4 - term5
    
    # Scale objective by J to prevent gradient collapse in finite-differencing
    return(-log_lik / J)
  }
  
  init_lambda <- 0.5
  init_log_nu0 <- log(5.0) 
  
  epsilon <- 1e-5 
  
  # nlminb is significantly more robust for this topology
  opt_res <- nlminb(
    start = c(init_lambda, init_log_nu0),
    objective = neg_log_pdf,
    lower = c(epsilon, log(epsilon)),
    upper = c(1 - epsilon, log(1e5))
  )
  
  return(list(
    lambda = opt_res$par[1],
    nu_0 = exp(opt_res$par[2]),
    # Scale the objective back up by J to get true log likelihood
    max_log_pdf = -opt_res$objective * J,
    convergence = opt_res$convergence, # 0 means successful
    message = opt_res$message
  ))
}
```

### 3.1.1. Moment estimate

```{r}
library(susieR)
fit_R_silly <- susie_suff_stat(
  XtX = N * R0,
  Xty = N * v,
  n = N,
  yty = N,
  L = 5,
  residual_variance = 1
)

b_bar = colSums(fit_R_silly$alpha * fit_R_silly$mu)


v0 = v - R0 %*% b_bar

lambda_est = moment_estimate_lambda(v0, R0, N)
lambda_est
```



```{r}
delta = 1
nu0 = delta * (1 - lambda_est) / lambda_est

nu0
```

```{r}
lambda <- lambda_est
diag_R <- diag(R)
k      <- min(N0, J)
nu_q = nu0 + 1

susie_love_pre_est_lambda <- susie_love_fix_2params(v, R0, diag_R,
                                   lambda, nu_q, nu_0 = nu0 + delta,
                                   N, L = 5,
                                   k,
                                   z_init = NULL,
                                   step_susie = 20,
                                   step_R = 30,
                                   step_nu = 50,
                                   max_iter = 500,
                                   max_iter_init = 100,
                                   tol = 1e-4)

```

```{r}
b_res <- susie_love_pre_est_lambda$b_res
R_q   <- susie_love_pre_est_lambda$R_q

susie_plot_alpha(
  alpha = b_res$alpha,
  R = cov2cor(R_q),
  b = beta_true,
  main = paste0("SuSiE-LovE pre-estimated lambda and nu0")
)
```

Hence, we achieve conservative behavior in this case using the moment estimate.
I actually find the moment estimate now too reliable because it is learn from `lm(w^2 ~ s)` and they look like this:

```{r}
eig = eigen(R0)
U = eig$vectors
s = eig$values
w = sqrt(N) * t(U) %*% v0
plot(s, w^2)
abline(lm(w^2 ~ s))
```


### 3.1.2. MLE of lambda

Let us try to use the MLE estimate of $\lambda$ to see if there is any difference.

```{r}
result <- maximize_log_pdf(R0, v0, N)
print(result)

lambda_est = result$lambda
lambda_est
```

It is interesting that the MLE of $\lambda$ is much lower than the moment estimate. Let's try susielove with this $\lambda$ and corresponding $\nu_0$.

```{r}
delta = 1
nu0 = delta * (1 - lambda_est) / lambda_est

nu0
```

```{r}
lambda <- lambda_est
diag_R <- diag(R)
k      <- min(N0, J)
nu_q = nu0 + 1

susie_love_pre_est_lambda <- susie_love_fix_2params(v, R0, diag_R,
                                   lambda, nu_q, nu_0 = nu0 + delta,
                                   N, L = 5,
                                   k,
                                   z_init = NULL,
                                   step_susie = 20,
                                   step_R = 30,
                                   step_nu = 50,
                                   max_iter = 500,
                                   max_iter_init = 100,
                                   tol = 1e-4)

b_res <- susie_love_pre_est_lambda$b_res
R_q   <- susie_love_pre_est_lambda$R_q

susie_plot_alpha(
  alpha = b_res$alpha,
  R = cov2cor(R_q),
  b = beta_true,
  main = paste0("SuSiE-LovE pre-estimated lambda and nu0 (MLE)")
)

```

I'd say that it is almost conservative for the MLE. Let me try a smarter $\nu_0$ set by the Wen and Stephens (2010):

$$\lambda \approx \dfrac{2\theta}{\theta + 2 \nu_0} \approx \dfrac{\theta}{\nu_0} \approx \dfrac{1}{\nu_0 (\log(2\nu_0) + 0.57)},$$
where $\theta \approx (\log(2 \nu_0) + 0.57)^{-1}$

```{r}
find_nu0_WS <- function(lambda) {
  f <- function(nu0) {
    1 / (nu0 * (log(2 * nu0) + 0.57)) - lambda
  }
  uniroot(f, interval = c(1, 1e6))$root
}

nu0_WS =  find_nu0_WS(lambda_est)
nu0_WS
```

Ok with this small nu0 we can expect conservative behavior.

```{r}
lambda <- lambda_est
diag_R <- diag(R)
k      <- min(N0, J)
nu_q = nu0_WS + 1

susie_love_pre_est_lambda <- susie_love_fix_2params(v, R0, diag_R,
                                   lambda, nu_q, nu_0 = nu0_WS,
                                   N, L = 5,
                                   k,
                                   z_init = NULL,
                                   step_susie = 20,
                                   step_R = 30,
                                   step_nu = 50,
                                   max_iter = 500,
                                   max_iter_init = 100,
                                   tol = 1e-4)

b_res <- susie_love_pre_est_lambda$b_res
R_q   <- susie_love_pre_est_lambda$R_q

susie_plot_alpha(
  alpha = b_res$alpha,
  R = cov2cor(R_q),
  b = beta_true,
  main = paste0("SuSiE-LovE pre-estimated lambda and nu0 (MLE + WS)")
)
```



### 3.2. But is it too conservative in other cases???

Now consider another case where the mismatch is low and the true $L = 2$. Let's see if it can recover both signals (and set $\nu_0$ to be large, $\lambda$ to be small) or it is still conservative. 

```{r}
seed  <- 1
T_split <- 1  ## divergence time very small so the mismatch is small
chrom <- "chr22"
start <- 20e6
end   <- 21e6
N     <- 25000L
N0    <- 1000L
J     <- 500L
h2    <- 0.01

res <- sim_2pop(chrom, start, end,
                T_split = T_split,
                n_gwas = N,
                n_ref = N0,
                J = J,
                h2 = h2,
                seed = seed)

v               <- res$v
R               <- res$R
R0              <- res$R0
beta_true       <- res$beta_true
causal_SNPs_loc <- res$causal_SNPs_loc

## standardized
d = 1 / sqrt(diag(R))
v = d * v
R = cov2cor(R)
R0 = cov2cor(R0)
  
plot(v, pch = 16)
points(causal_SNPs_loc, v[causal_SNPs_loc], pch = 16, col = "red")
```

Here is susie fitted with the out-of-sample LD:

```{r}
fit_R_init <- susie_suff_stat(
  XtX = N * R0,
  Xty = N * v,
  n = N,
  yty = N,
  L = 5,
  residual_variance = 1
)

susie_plot(fit_R_init, y = "PIP", b = beta_true,
           main = "Out-of-sample corr")
```

Even the out-of-sample LD works well. We know that the mismatch is small.

```{r}
b_bar = colSums(fit_R_init$alpha * fit_R_init$mu)


v0 = v - R0 %*% b_bar

lambda_est = moment_estimate_lambda(v0, R0, N)
lambda_est
```

I do not understand why estimated $\lambda$ is large. I decided to print the $w^2_j$ and $s_j$ inside the moment estimating formula to see what the linear model is fitting to.

```{r}
eig = eigen(R0)
U = eig$vectors
s = eig$values
w = sqrt(N) * t(U) %*% v0
plot(s, w^2)
abline(lm(w^2 ~ s))
```

Ok no wonder why it is bad. Let's try MLE next.

### 3.2.2. MLE

```{r}
result <- maximize_log_pdf(R0, v0, N)
print(result)

lambda_est = result$lambda
lambda_est
```

Learning $\lambda$ using MLE is surprisingly different. $\lambda$ here is very small, which is expected for small mismatch.

```{r}
nu0_WS =  find_nu0_WS(lambda_est)
nu0_WS
```

susie-love result:

```{r}
lambda <- lambda_est
diag_R <- diag(R)
k      <- min(N0, J)
nu_q = nu0_WS + 1

susie_love_pre_est_lambda <- susie_love_fix_2params(v, R0, diag_R,
                                   lambda, nu_q, nu_0 = nu0_WS,
                                   N, L = 5,
                                   k,
                                   z_init = NULL,
                                   step_susie = 20,
                                   step_R = 30,
                                   step_nu = 50,
                                   max_iter = 500,
                                   max_iter_init = 100,
                                   tol = 1e-4)

b_res <- susie_love_pre_est_lambda$b_res
R_q   <- susie_love_pre_est_lambda$R_q

susie_plot_alpha(
  alpha = b_res$alpha,
  R = cov2cor(R_q),
  b = beta_true,
  main = paste0("SuSiE-LovE pre-estimated lambda and nu0 (MLE + WS)")
)
```

OKKKKKKK!!!!!!!!

