---
title: "Learning susie-love with various lambda"
Author: "Dat Do"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
---

I am sorry for saying that learned $\lambda$ is either 0 or 1. I missed the logdet
term in the objective function when trying to think about it. 
My frustration comes from an earlier moment
when I attempted to learn both $\lambda$ and $\nu_0$. 
I set the upper bound of $\lambda$ to be 0.1. The learned $\lambda$ just kept 
increasing to this upper bound
as the loss function decreases (and $\nu_0$ also increases). This makes the 
algorithm non-robust (because we borrow too much information from LD with 
large $\lambda$ -> wrong LD). 

Although the current implement fixes $\lambda$, we will demonstrate here that
changing $\lambda$ to a larger value gives better loss (loss = - ELBO) and less
robust result. 

Note that $R$ and $R0$ in this markdown file are covariance matrices (but
we abuse the language/notation and refer them as LD).

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 1. Simulate data from a coalescent model

First we use `sim_2pop()` to simulate two populations that diverged 25 generations
ago. POP1 is a large GWAS cohort (N = 25,000) and POP2 is a small LD
reference panel (N₀ = 200).

```{r simulate, fig.width=8, fig.height=6}
library(susielove)

seed  <- 1
T_split <- 25
chrom <- "chr22"
start <- 20e6
end   <- 21e6
N     <- 25000L
N0    <- 200L
J     <- 1000L
h2    <- 0.01

res <- sim_2pop(chrom, start, end,
                T_split = T_split,
                n_gwas = N,
                n_ref = N0,
                J = J,
                h2 = h2,
                seed = seed)

v               <- res$v
R               <- res$R
R0              <- res$R0
beta_true       <- res$beta_true
causal_SNPs_loc <- res$causal_SNPs_loc

plot(v, pch = 16)
points(causal_SNPs_loc, v[causal_SNPs_loc], pch = 16, col = "red")
```

## 2. Fit SuSiE with in-sample covariance matrix (oracle baseline)

When the true in-sample LD is available, standard SuSiE works perfectly:

```{r susie-insample, fig.width=8, fig.height=6}
library(susieR)

fit <- susie_suff_stat(
  XtX = N * R,
  Xty = N * v,
  n = N,
  yty = N
)
susie_plot(fit, y = "PIP", b = beta_true, main = "In-sample cov mat")
```

## 3. Fit SuSiE with out-of-sample covariance matrix

Using the mismatched LD reference directly leads to spurious signals:

```{r susie-outsample, fig.width=8, fig.height=6}
fit_R0 <- susie_suff_stat(
  XtX = N * R0,
  Xty = N * v,
  n = N,
  yty = N,
  L = 5
)
susie_plot(fit_R0, y = "PIP", b = beta_true, main = "Out-of-sample cov mat")
```

## 4. Diagonal correction

Here we use Matthew's suggestion that we can adjust the diagonal of the out-of-sample
to the diagonal of the in-sample. It works a bit better.

```{r diag-correction, fig.width=8, fig.height=6}
d0_inv_sqrt <- 1 / sqrt(diag(R0))
LD0 <- diag(d0_inv_sqrt) %*% R0 %*% diag(d0_inv_sqrt)
d_sqrt <- sqrt(diag(R))
R_tilde <- diag(d_sqrt) %*% LD0 %*% diag(d_sqrt)

fit_R_tilde <- susie_suff_stat(
  XtX = N * R_tilde,
  Xty = N * v,
  n = N,
  yty = N
)
susie_plot(fit_R_tilde, y = "PIP", b = beta_true,
           main = "Out-of-sample cov mat (corrected diag)")
```

## 5. Fit SuSiE-LovE

`susie_love_lambda` jointly learns the covariance matrix while performing
fine-mapping. Here $\lambda$ is set to $1 / (N_0 + 1) \approx 1 / 200 = 0.005$

```{r susie-love, fig.width=8, fig.height=6}

lambda <- 1 / (N0 + 1)
diag_R <- diag(R)
R_out  <- R_tilde
k      <- min(N0, J)

love_res <- susie_love_lambda(v, R_out, diag_R,
                              lambda,
                              N, L = 5,
                              k,
                              nu_0_init = 1000,
                              step_susie = 20,
                              step_R = 30,
                              step_nu = 10,
                              max_iter = 30, 
                              tol = 1e-4)

b_res <- love_res$b_res
R_q   <- love_res$R_q

susie_plot_alpha(
  alpha = b_res$alpha,
  R = cov2cor(R_q),
  b = beta_true,
  main = paste0("SuSiE-LovE with lambda = ", lambda)
)
```
The result looks good. The spurious CS is gone. 

## 6. Fit susie-love with higher lambda

We attempt to fit the susie-love model with $\lambda = 0.02$ (4 times larger than
above). We observe that the loss is smaller than the fit above. $\nu_0$ is much
larger. And we might have spurious CS (or at least spurious non-zero PIP for many SNPs) again.

```{r lambda 002, fig.width=8, fig.height=6}
lambda <- 0.02
diag_R <- diag(R)
R_out  <- R_tilde
k      <- min(N0, J)

love_res <- susie_love_lambda(v, R_out, diag_R,
                              lambda,
                              N, L = 5,
                              k,
                              nu_0_init = 1000,
                              step_susie = 20,
                              step_R = 30,
                              step_nu = 10,
                              max_iter = 30, 
                              tol = 1e-4)

b_res <- love_res$b_res
R_q   <- love_res$R_q

susie_plot_alpha(
  alpha = b_res$alpha,
  R = cov2cor(R_q),
  b = beta_true,
  main = paste0("SuSiE-LovE with lambda = ", lambda)
)
```
Now we set $\lambda = 0.05$, even higher. We often have better loss. 

```{r lambda 005, fig.width=8, fig.height=6}
lambda <- 0.05
diag_R <- diag(R)
R_out  <- R_tilde
k      <- min(N0, J)

love_res <- susie_love_lambda(v, R_out, diag_R,
                              lambda,
                              N, L = 5,
                              k,
                              nu_0_init = 1000,
                              step_susie = 20,
                              step_R = 30,
                              step_nu = 10,
                              max_iter = 30, 
                              tol = 1e-4)

b_res <- love_res$b_res
R_q   <- love_res$R_q

susie_plot_alpha(
  alpha = b_res$alpha,
  R = cov2cor(R_q),
  b = beta_true,
  main = paste0("SuSiE-LovE with lambda = ", lambda)
)
```

The loss does not increase further if we keep increase $\lambda$. But the learned
$\nu_0$ is still large compare to the case where we set $\lambda = 1 / (N_0 + 1)$.

```{r lambda 01, fig.width=8, fig.height=6}
lambda <- 0.1
diag_R <- diag(R)
R_out  <- R_tilde
k      <- min(N0, J)

love_res <- susie_love_lambda(v, R_out, diag_R,
                              lambda,
                              N, L = 5,
                              k,
                              nu_0_init = 1000,
                              step_susie = 20,
                              step_R = 30,
                              step_nu = 10,
                              max_iter = 30, 
                              tol = 1e-4)

b_res <- love_res$b_res
R_q   <- love_res$R_q

susie_plot_alpha(
  alpha = b_res$alpha,
  R = cov2cor(R_q),
  b = beta_true,
  main = paste0("SuSiE-LovE with lambda = ", lambda)
)
```


