---
title: "Setting small nu leads to conservative behaviour"
Author: "Dat Do"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
editor_options: 
  markdown: 
    wrap: 72
---

## 1. Simulate data from a coalescent model

```{r simulate, fig.width=8, fig.height=6}
library(susielove)

seed  <- 1
T_split <- 25
chrom <- "chr22"
start <- 20e6
end   <- 21e6
N     <- 25000L
N0    <- 200L
J     <- 1000L
h2    <- 0.01

res <- sim_2pop(chrom, start, end,
                T_split = T_split,
                n_gwas = N,
                n_ref = N0,
                J = J,
                h2 = h2,
                seed = seed)

v               <- res$v
R               <- res$R
R0              <- res$R0
beta_true       <- res$beta_true
causal_SNPs_loc <- res$causal_SNPs_loc
true_SNP = which(beta_true != 0)

plot(v, pch = 16)
points(causal_SNPs_loc, v[causal_SNPs_loc], pch = 16, col = "red")
```

## 2. SuSiE with out-of-sample LD:

```{r diag-correction, fig.width=8, fig.height=6}
library(susieR)
d0_inv_sqrt <- 1 / sqrt(diag(R0))
LD0 <- diag(d0_inv_sqrt) %*% R0 %*% diag(d0_inv_sqrt)
d_sqrt <- sqrt(diag(R))
R_tilde <- diag(d_sqrt) %*% LD0 %*% diag(d_sqrt)

fit_R_tilde <- susie_suff_stat(
  XtX = N * R_tilde,
  Xty = N * v,
  n = N,
  yty = N
)
susie_plot(fit_R_tilde, y = "PIP", b = beta_true,
           main = "Out-of-sample cov mat (corrected diag)")
```

## 3. SuSiE-Love with leared nu0

```{r susie-love, fig.width=8, fig.height=6}

lambda <- 1 / (N0 + 1)
diag_R <- diag(R)
R_out  <- R_tilde
k      <- min(N0, J)

love_res <- susie_love_lambda(v, R_out, diag_R,
                              lambda,
                              N, L = 5,
                              k,
                              nu_0_init = 1000,
                              step_susie = 20,
                              step_R = 30,
                              step_nu = 10,
                              max_iter = 30, 
                              tol = 1e-4)

b_res <- love_res$b_res
R_q   <- love_res$R_q

susie_plot_alpha(
  alpha = b_res$alpha,
  R = cov2cor(R_q),
  b = beta_true,
  main = paste0("SuSiE-LovE with lambda = ", lambda)
)
```

## 4. SuSiE-LovE with fixed and small nu0

Now let's try to set $\nu_0$ to very small to see if it can be more
conservative:

```{r}
R_out  <- R_tilde
## nu_0 = a * N_0 + delta, and we set delta = 1, so minimal nu_0 = 1
nu_0 = 1.1  
lambda <- 1 / nu_0
diag_R <- diag(R)
k      <- min(N0, J)
nu_q = nu_0 + 1

susie_love_small_nu <- susie_love_fix_2params(v, R_out, diag_R,
                                   lambda, nu_q, nu_0,
                                   N, L = 5,
                                   k,
                                   z_init = NULL,
                                   step_susie = 20,
                                   step_R = 30,
                                   step_nu = 50,
                                   max_iter = 500,
                                   max_iter_init = 100,
                                   tol = 1e-4)

b_res <- susie_love_small_nu$b_res
R_q   <- susie_love_small_nu$R_q

susie_plot_alpha(
  alpha = b_res$alpha,
  R = cov2cor(R_q),
  b = beta_true,
  main = paste0("SuSiE-LovE with small nu0 = ", nu_0)
)
```

The algorithm needs more iterations to converge. But it does lead to L =
1.

## 5. Investigate Rq (the estimated covariance matrix in susie-love having conservative behaviour)

Now let's check if the learned LD is similar to our guess (discussion
with Matthew):

$$R_{null} = S + \eta vv^\top, $$ where $\eta$ is chosen to be as large
as possible, and $S$ is a digonal matrix so that the diagonal of this
$R_{null}$ equals to the diagonal of the in-sample LD matrix $\diag(R)$.

```{r}
eta = min(diag_R / v^2)
## check
diag_diff = diag_R - diag(eta * tcrossprod(v))
min(diag_diff)  ## should = 0

R_null = eta * tcrossprod(v)
diag(R_null) = diag_R
zlim <- range(R_null, R_q)

par(mfrow = c(1, 2))

## first 100 x 100 values
image(R_null[1:100, 1:100], main = "Null cov mat", col = heat.colors(20), zlim = zlim)
image(R_q[1:100, 1:100], main = "Learned cov mat", col = heat.colors(20), zlim = zlim)
par(mfrow = c(1, 1))
```

Ok they don't look similar at all. Let check the rank of the learned
matrix

```{r}
eig = eigen(R_q)
envals = eig$values

plot(c(1:J), envals, pch = 16)
print(envals[c(1:10)])
```

Indeed $R_q$ has rank close to 1. Rank of the guessed null matrix:

```{r}
eig_null = eigen(R_null)
envals_null = eig_null$values

plot(c(1:J), envals_null, pch = 16)
print(envals_null[c(1:10)])
```

It is also close to 1. I am still investigate to see what is the form of
this matrix $R_q$.

## More visualization of the learned Rq

### Row of Rq at the true SNP

I visualize the value of Rq at the row of the true causal SNP that it
catches. Indeed the value is very close to $v$.

```{r}
plot(c(1:J), R_q[true_SNP[1], ] * beta_true[true_SNP[1]], col = 'red')
points(c(1:J), v, col='blue')
```

### Other rows

For all other rows it has this weird behaviour:

```{r}
plot(c(1:J), R_q[1, ])

```

```{r}
plot(c(1:J), R_q[2, ], col = 'black')
points(c(1:J), R_q[3, ] / R_q[3, 1] * R_q[2, 1], col = 'red')

```

```{r}
plot(c(1:J), R_q[2, ], col = 'black')
points(c(1:J), R_q[3, ] / R_q[3, 1] * R_q[2, 1], col = 'red')
```

The R_null that we proposed before has all rows similar to v (in each
row, the biggest value is the same as the corresponding value on the
diagonal of R).

```{r}
plot(c(1:J), R_null[1, ])
which.max(R_null[1, ])
```

```{r}
plot(c(1:J), R_null[2, ])
which.max(R_null[2, ])
```

### Conclusion on the Rq

Because except the row/column corresponding to one of the true causal
SNP being similar to v, all the other rows of $R_q$ are multiplier of
others, we can conclude that $R_q$ has this form:
$$R_q = u u^\top + \eta vv^\top, $$ where $u \in \mathbb{R}^J$. Our
intuition of R_null is like a diagonal matrix + vvT. But perhaps R_null
is more like rank-one + vvT.

Anyway, this R_q still has the effect that $v \approx R_q b_1$, where
$b_1$ is a single effect vector (i.e., after taking this single effect
out, there is no left signal to learn the second single effect).

```{r}
b_bar = colSums(b_res$alpha * b_res$mu)
leftover_signal = v - R_q %*% b_bar
t(leftover_signal)[1:100]
```

One simple way to make inference about $\eta$ (the multiplier of
$vv^\top$) in $R_q$ is as follows:

```{r}
eta = as.numeric(1 / v %*% solve(R_q + 1e-10 * diag(J), v))
eta
```

After having this $\eta$, the vector $u = (u_j)_{j=1}^{J}$ is
constructed so that
$$\text{diag}(u u^\top + \eta vv^\top) = \text{diag}(R)$$ in other
words, $$u_j^2 = R_{jj} - \eta v_j^2, \quad \forall j,$$ and the signs
of each $u_j$ can be arbitrary.

```{r}
uuT = R_q - eta * tcrossprod(v)
```

# Conclusion

The conservative covariance matrix that we get from susie-love is
approximately of the form rank-one-plus-vvT:
$$R_{conservative} = uu^\top + \eta vv^\top, $$ Here is a theoretical
characterization of $u$ and $\eta$:

1.  Firstly, because $\eta$ capture the maximmal information of
    $vv^\top$ in $R_{conservative}$, we have
    $$v^\top (R_{conservative} - \eta vv^\top) v = 0$$ which leads to
    $(v^\top u)^2 = 0$. This implies $u \perp v$.

2.  Suppose that the learned $\bar{b} = \beta_1 \gamma_1$ with
    $\gamma_1 = e_k$ has 1 at $k$ and 0 otherwise and
    $\beta_1 \in \mathbb{R}$. The fact that (I change $\approx$ to $=$ for simplicity)
    $$v = R_{conservative} \overline{b} $$ leads to
    $$v = u u^\top + \eta vv^\top (\beta_1 e_k) = \beta_1 u_k u + \beta_1 \eta v_k v$$
implies $u_k = 0$ (since $u\perp v$) and $\eta = 1 / (\beta_1 * v_k)$. 

We can check this

```{r}
u = uuT[1, ] / sqrt(abs(uuT[1, 1]))
u %*% v    ## should near 0; subject to the error due to the low-rank part (R_conservative is still a full rank matrix anyway)
```

```{r}
k = true_SNP[1]
u[k] 
```

This u_k is near 0 as it is significantly smaller than other u

```{r}
u[1]
u[2]
u[3]
```

Finally we check the identity of $\eta$:

```{r}
eta * v[k] * beta_true[k]  ## should near 1
```
