---
title: "Learning lambda and nu0 in some special settings"
Author: "Dat Do"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
---

# 1. MLE in the null case
Consider the model
$$v | R \sim \mathcal{N}(0, R/N), \quad R | R_0 \sim \mathcal{IW}(\nu_0 \overline{R}, \nu_0 + J + 1), $$
for $\overline{R} = (1 - \lambda) R_0 + \lambda I$. The following code finds the MLE of $\nu_0$ and $\lambda$.

```{r}
maximize_log_pdf <- function(R0, v, N) {
  J <- length(v)
  
  R0 <- (R0 + t(R0)) / 2
  eig <- eigen(R0, symmetric = TRUE)
  
  # Clamp any floating-point negative eigenvalues to exactly 0
  d <- pmax(eig$values, 0) 
  v_tilde <- as.numeric(crossprod(eig$vectors, v))
  
  # Optimize over log_nu0 to keep parameter scales identical
  neg_log_pdf <- function(params) {
    lambda <- params[1]
    nu0 <- exp(params[2]) # Decode log(nu0)
    
    d_lambda <- (1 - lambda) * d + lambda
    
    log_det <- sum(log(d_lambda))
    qf <- N * sum((v_tilde^2) / d_lambda)
    
    inner_log <- 1 + (1 / nu0) * qf
    
    term1 <- lgamma((nu0 + J + 2) / 2)
    term2 <- lgamma((nu0 + 2) / 2)
    term3 <- (J / 2) * log(nu0 * pi)
    term4 <- 0.5 * log_det
    term5 <- ((nu0 + J + 2) / 2) * log(inner_log)
    
    log_lik <- term1 - term2 - term3 - term4 - term5
    
    # Scale objective by J to prevent gradient collapse in finite-differencing
    return(-log_lik / J)
  }
  
  init_lambda <- 0.5
  init_log_nu0 <- log(5.0) 
  
  epsilon <- 1e-5 
  
  # nlminb is significantly more robust for this topology
  opt_res <- nlminb(
    start = c(init_lambda, init_log_nu0),
    objective = neg_log_pdf,
    lower = c(epsilon, log(epsilon)),
    upper = c(1 - epsilon, log(1e5))
  )
  
  return(list(
    lambda = opt_res$par[1],
    nu_0 = exp(opt_res$par[2]),
    # Scale the objective back up by J to get true log likelihood
    max_log_pdf = -opt_res$objective * J,
    convergence = opt_res$convergence, # 0 means successful
    message = opt_res$message
  ))
}
```

Now we simulate under this exact model with some true $\lambda$ and $\nu_0$ to ensure that $\lambda$ can be learned correctly. Estimation of $\nu_0$ is off.

```{r}
# ---- 1. True model simulation check
J <- 1000
set.seed(1)
N = 1000
lambda = 0.01
nu0 = 100

## sample R0
X <- matrix(rnorm(N * J), N, J)
X_std <- scale(X)
R0 <- cov(X_std)
R_bar = (1 - lambda) * R0 + lambda * diag(J) 

## sample R
library(MCMCpack)
R <- riwish(v = nu0 + J + 1, S = nu0 * R_bar)

## sample v
diag(R) <- diag(R) + 1e-10  ##safety
L <- t(chol(R))
z <- rnorm(J)
v <- as.numeric(L %*% z) / sqrt(N)

result <- maximize_log_pdf(R0, v, N)
print(result)
```

Change the true $\lambda$: The MLE still can learn the correct $\lambda$.
```{r}
J <- 1000
set.seed(1)
N = 1000
lambda = 0.02
nu0 = 100

## sample R0
X <- matrix(rnorm(N * J), N, J)
X_std <- scale(X)
R0 <- cov(X_std)
R_bar = (1 - lambda) * R0 + lambda * diag(J) 

## sample R
library(MCMCpack)
R <- riwish(v = nu0 + J + 1, S = nu0 * R_bar)

## sample v
diag(R) <- diag(R) + 1e-10  ##safety
L <- t(chol(R))
z <- rnorm(J)
v <- as.numeric(L %*% z) / sqrt(N)

result <- maximize_log_pdf(R0, v, N)
print(result)
```

Now we try to learn $\lambda$ and $\nu_0$ from a null observation from coalescent simulation (note that in this case we do not have true $\lambda$ and $\nu_0$, but it helps with the intuition on how large $\lambda$ should be in this case).

First we try with the out-of-sample size being 200. We see that the optimal $\lambda = 0.05$ when $N_0 = 200$.


```{r}
library(susielove)
N = 5000
J = 1000
ret = sim_2pop(chrom = "chr22",
               start = 20e6,
               end = 21e6,
               T_split = 25,
               n_gwas = N,
               n_ref = 200,
               J = J,
               h2 = .01,
               seed = 1)

R0 = ret$R0
R = ret$R

R0 = cov2cor(R0)
R = cov2cor(R)

diag(R) <- diag(R) + 1e-10  ##safety
L <- t(chol(R))
z <- rnorm(J)
v <- as.numeric(L %*% z) / sqrt(N)

result <- maximize_log_pdf(R0, v, N)
print(result)
```
Next we try with the out-of-sample size being 2000. We see that the optimal $\lambda = 0.0001$ when $N_0 = 2000$. Both cases have very small $\nu_0$ though.

```{r}
library(susielove)
N = 5000
J = 1000
ret = sim_2pop(chrom = "chr22",
               start = 20e6,
               end = 21e6,
               T_split = 25,
               n_gwas = N,
               n_ref = 2000,
               J = J,
               h2 = .01,
               seed = 1)

R0 = ret$R0
R = ret$R

R0 = cov2cor(R0)
R = cov2cor(R)

diag(R) <- diag(R) + 1e-10  ##safety
L <- t(chol(R))
z <- rnorm(J)
v <- as.numeric(L %*% z) / sqrt(N)

result <- maximize_log_pdf(R0, v, N)
print(result)

```

# 2. MLE in the single effect case (without noise)

Here we implement the MLE of the setting assuming that we know the single true effect and $N\to \infty$ (so that the variance of $v$ is 0). This is equivalent to observing the first column of $R$. The formula appears in Section 2 of the new small writeup.

```{r}
optimize_params_single_effect <- function(R_col1, R0, init_nu0 = 50, init_lambda = 0.5) {
  
  # Length of the column vector (p dimension)
  p <- length(R_col1)
  
  # Identity matrix of dimension p
  I_p <- diag(p)
  
  # Partition the observed first column of R
  r_11 <- R_col1[1]
  R_21 <- R_col1[-1]
  
  # ---------------------------------------------------------
  # Objective Function (Negative Log-Likelihood)
  # ---------------------------------------------------------
  neg_log_lik <- function(params) {
    nu_0   <- params[1]
    lambda <- params[2]
    
    # 1. Dynamically construct Rbar based on the current lambda
    Rbar <- (1 - lambda) * R0 + lambda * I_p
    
    # 2. Partition the dynamic Rbar matrix
    rbar_11 <- Rbar[1, 1]
    Rbar_21 <- Rbar[-1, 1, drop = FALSE] # Keep as a column matrix
    Rbar_22 <- Rbar[-1, -1, drop = FALSE]
    
    # 3. Compute the Schur Complement of Rbar dynamically
    # Rbar_22.1 = Rbar_22 - Rbar_21 * (1/rbar_11) * Rbar_12
    Rbar_22.1 <- Rbar_22 - (Rbar_21 %*% t(Rbar_21)) / rbar_11
    
    # Compute inverse and log-determinant (safely)
    inv_Rbar_22.1 <- solve(Rbar_22.1)
    log_det_Rbar_22.1 <- as.numeric(determinant(Rbar_22.1, logarithm = TRUE)$modulus)
    
    # --- A. Log-Likelihood of r_11 (Inverse-Gamma) ---
    alpha <- (nu_0 + 2) / 2
    beta  <- (nu_0 * rbar_11) / 2
    
    ll_r11 <- alpha * log(beta) - lgamma(alpha) - 
      (alpha + 1) * log(r_11) - (beta / r_11)
    
    # --- B. Log-Likelihood of R_21 | r_11 (Multivariate t) ---
    mu <- Rbar_21 * (r_11 / rbar_11)
    diff <- R_21 - mu
    
    # Quadratic form: (R_21 - mu)^T %*% Rbar_22.1^{-1} %*% (R_21 - mu)
    Q <- as.numeric(t(diff) %*% inv_Rbar_22.1 %*% diff)
    
    term1 <- lgamma((nu_0 + p + 2) / 2) - lgamma((nu_0 + 3) / 2)
    term2 <- - ((p - 1) / 2) * log(pi)
    term3 <- - ((p - 1) / 2) * log(r_11 / rbar_11)
    term4 <- - 0.5 * log_det_Rbar_22.1
    term5 <- - ((nu_0 + p + 2) / 2) * log(1 + (rbar_11 / r_11) * Q)
    
    ll_R21_given_r11 <- term1 + term2 + term3 + term4 + term5
    
    # Return total negative log-likelihood
    return(-(ll_r11 + ll_R21_given_r11))
  }
  
  # ---------------------------------------------------------
  # Execution of the Optimizer
  # ---------------------------------------------------------
  # "L-BFGS-B" allows us to set both lower and upper bounds
  # nu_0 must be > 0. lambda must be in [0, 1].
  opt_result <- optim(
    par       = c(init_nu0, init_lambda),
    fn        = neg_log_lik,
    method    = "L-BFGS-B",
    lower     = c(1e-4, 1e-4),    # Keep strictly > 0
    upper     = c(Inf, 0.9999),   # lambda caps at 1
    control   = list(maxit = 10000)
  )
  
  return(list(
    nu_0_opt    = opt_result$par[1],
    lambda_opt  = opt_result$par[2],
    neg_log_lik = opt_result$value,
    convergence = opt_result$convergence,
    message     = opt_result$message
  ))
}
```

We first simulate under the correct model to check the estimation of $\lambda$ and $\nu_0$. It is awesome that the MLE can learn both of them pretty accurately. The estimation of $\nu_0$ is way better compared to the null case.

```{r}
# ---- 1. True model simulation check
J <- 1000
set.seed(1)
N = 1000
lambda = 0.01
nu0 = 400

## sample R0
X <- matrix(rnorm(N * J), N, J)
X_std <- scale(X)
R0 <- cov(X_std)
R_bar = (1 - lambda) * R0 + lambda * diag(J) 

## sample R
library(MCMCpack)
R <- riwish(v = nu0 + J + 1, S = nu0 * R_bar)
R_first_col = R[1, ]

results <- optimize_params_single_effect(
  R_col1      = R_first_col, 
  R0          = R0, 
  init_nu0    = 50, 
  init_lambda = 0.5
)

print(results)
```
Change the true $\lambda$ and $\nu_0$ to see that we still have good results. (Idk why estimation of $\lambda$ is off in this case tho. $\nu_0$ is good.)

```{r}
J <- 1000
set.seed(1)
N = 10000
lambda = 0.03
nu0 = 1000

## sample R0
X <- matrix(rnorm(N * J), N, J)
X_std <- scale(X)
R0 <- cov(X_std)
R_bar = (1 - lambda) * R0 + lambda * diag(J) 

## sample R
library(MCMCpack)
R <- riwish(v = nu0 + J + 1, S = nu0 * R_bar)
R_first_col = R[1, ]

results <- optimize_params_single_effect(
  R_col1      = R_first_col, 
  R0          = R0, 
  init_nu0    = 1000, 
  init_lambda = 0.5
)

print(results)
```



Now we turn into coalescent simulation to get an idea about the estimation of $\nu_0$ and $\lambda$. First try with $N_0 = 200$. This leads to similar $\lambda$ to the null case, but will a larger $\nu_0$.

We try with larger $N_0$, which leads to smaller $\lambda$ and larger $\nu_0$.

```{r}

# ---- 2. Coalescent simulation
library(susielove)
N = 5000
J = 1000
ret = sim_2pop(chrom = "chr22",
               start = 20e6,
               end = 21e6,
               T_split = 25,
               n_gwas = N,
               n_ref = 200,
               J = J,
               h2 = .01,
               seed = 1)

R0 = ret$R0
R = ret$R

R0 = cov2cor(R0)
R = cov2cor(R)

R_first_col = R[1, ]

results <- optimize_params_single_effect(
  R_col1      = R_first_col, 
  R0          = R0, 
  init_nu0    = 100, 
  init_lambda = 0.1
)

print(results)
```
```{r}
library(susielove)
N = 5000
J = 1000
ret = sim_2pop(chrom = "chr22",
               start = 20e6,
               end = 21e6,
               T_split = 25,
               n_gwas = N,
               n_ref = 2000,
               J = J,
               h2 = .01,
               seed = 1)

R0 = ret$R0
R = ret$R

R0 = cov2cor(R0)
R = cov2cor(R)

R_first_col = R[1, ]

results <- optimize_params_single_effect(
  R_col1      = R_first_col, 
  R0          = R0, 
  init_nu0    = 100, 
  init_lambda = 0.1
)

print(results)

```

