---
title: "Systematic comparison of population correlation matrix shrinkage method"
author: "dodat-stats"
date: "2026-06-04"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 1. Aim, and Recap the shrinkage estimate of the out-of-sample population correlation matrix

In this workflow file, we will compare the three shrinkage method for the out-of-sample population correlation matrix $\Psi_0$ more systematically (by comparing their corresponding `susie_love_lambda` result). Recall that we had 3 ways (naive, Wen and Stephens, and Ledoit and Wolf) to choose $\lambda$ in
$$\Psi_0 = (1 - \lambda) R_0 + \lambda I, $$
and then in susie-love we model
$$R | \nu_0 \sim \mathcal{IW}(\nu_0 \Psi_0, \nu_0 + J + 1).$$
(I will replace $\nu_0$ by $\eta_0$ when we revise the manuscript.) The reasons we are doing this are:

**First reason (phylosophical reason):** De-couple the roles of $\lambda$ and $\nu_0$: 

(i) Now $\lambda$ only helps the out-of-sample $R_0$ to become $\Psi_0$, the population version with better condition number (no more rank-deficient). $\lambda$ does not serve the role of quantifying the LD mismatch anymore [Therefore its role is now different from what was discussed in Zou et al. (2022)]. 

(ii) After obtaining $\Psi_0$, now $\nu_0$ has only one job which focuses on quantifying the LD mismatch between $R$ and $\Psi_0$.  

**Second reason (technical):** Jointly learning $\lambda$ (and also $\delta$ in the other version) inside susie_love is hard... Because we jointly model $R$ using the Inverse-Wishart. The log-det term in the log-likelihood really prefers matrices with moderate-large eigenvalues, increasing $\lambda$ to be higher than we want. Therefore I decided to just fix $\lambda$ in the beginning and only care about $\nu_0$. Luckily, there are principal ways to find $\lambda$ in the literature, such as in Leidoit and Wolf (2004) where they aim to minimize the Frobenius norm loss between the sample covariance matrix and population version (they a 2020 review paper that explains this clearly and enjoy it a lot), or in Wen and Stephens (2010) where they introduced a shrinkage estimate motivated from Li and Stephens (2003). Note that because we are working in a local part of the genome, we can ignore the recombination rate $\exp(-\rho / d_{ij})$ factor.

Recall 3 shrinkage (choosing $\lambda$) rules:

1. Naive shrinkage: $\Psi_{naive} = \left(1-\dfrac{1}{N_0}) R_0 + \dfrac{1}{N_0} I$

2. Wen and Stephens (2010) shrinkage: $\Psi_{WS} = \left(1- \lambda)^2 R_0 + (1 - (1-\lambda)^2) I$, where $\lambda = \dfrac{\theta}{\theta + 2 N_0}$ and $\theta$ is the Watterson's estimator $\theta = \left(\sum_{i=1}^{2N_0 - 1} \dfrac{1}{i}\right)^{-1} \approx (\log(2N_0) + 0.57)^{-1} $.

3. Ledoit and Wolf (2004) linear shrinkage: $\Psi_{linear-LW} = \left(1- \lambda) R_0 + \lambda I$, where $\lambda \asymp \dfrac{1}{N_0}$. 


```{r}


compute_wen_stephens <- function(R_0, N_ref) {
  # Number of haplotypes (diploid population)
  d <- 2 * N_ref
  
  # Calculate Watterson's estimator constant (Harmonic sum)
  harmonic_sum <- sum(1 / (1:(d - 1)))
  theta        <- 1 / harmonic_sum
  
  # Calculate the exact mutation shrinkage parameter lambda
  lambda <- theta / (theta + 2 * N_ref)
  
  # Apply local uniform shrinkage: Psi = (1-lambda)^2 * R_0 + (1 - (1-lambda)^2) * I
  p <- ncol(R_0)
  R_ws <- (1 - lambda)^2 * R_0 + (1 - (1 - lambda)^2) * diag(p)
  
  return(R_ws)
}


compute_lw_linear_from_R0 <- function(R_0, N_ref) {
  p <- ncol(R_0)
  
  # 1. Sum of the squared off-diagonal elements of R_0
  # (This is the Frobenius norm of the off-diagonals)
  sum_sq_off_diag <- sum(R_0^2) - p
  
  # 2. Estimate the sampling variance (noise) parameter under normality
  # In correlation space, the sum of variances of r_ij is approximated by:
  beta_sq <- (1 / N_ref) * (p + sum_sq_off_diag)
  
  # 3. Calculate the distance to the identity target matrix
  # Since diag(R_0) is all 1s and diag(I) is all 1s, the distance is just 
  # the sum of the squared off-diagonal elements.
  delta_sq <- sum_sq_off_diag
  
  # 4. Compute the optimal shrinkage intensity alpha
  # Guard against division by zero if R_0 is already the identity matrix
  if (delta_sq == 0) {
    alpha <- 1
  } else {
    alpha <- min(1, beta_sq / delta_sq)
  }
  
  # 5. Construct the shrunk matrix: alpha * I + (1 - alpha) * R_0
  R_lw_linear <- (1 - alpha) * R_0 + alpha * diag(p)
  
  # Print the learned alpha for transparency
  cat("Ledoit-Wolf Linear Alpha estimated from R_0:", alpha, "\n")
  
  return(R_lw_linear)
}


```


# 1. Running and comparing 3 methods on many simulation

See the end of the result.

```{r simulate}
library(susielove)
library(susieR)

num_reps = 25
coverages = matrix(0, nrow = num_reps, ncol = 5)
powers = matrix(0, nrow = num_reps, ncol = 5)
no_CSs = matrix(0, nrow = num_reps, ncol = 5)
no_SNPs_CS = matrix(0, nrow = num_reps, ncol = 5)
dist_R = matrix(0, nrow = num_reps, ncol = 4)


T_split <- 25
chrom <- "chr22"
start <- 20e6
end   <- 21e6
N     <- 25000L
N0    <- 200L
J     <- 1000L
h2    <- 0.01

for (seed in c(1:num_reps)){
  print(paste0("Seed ", seed))
  
  res <- sim_2pop(chrom, start, end,
                T_split = T_split,
                n_gwas = N,
                n_ref = N0,
                J = J,
                h2 = h2,
                seed = seed)

  v               <- res$v
  R               <- res$R
  R0              <- res$R0
  beta_true       <- res$beta_true
  causal_SNPs_loc <- res$causal_SNPs_loc
  
  ## standardized
  d = 1 / sqrt(diag(R))
  v = d * v
  R = cov2cor(R)
  R0 = cov2cor(R0)
  
  ## 1. in-sample LD
  fit <- susie_suff_stat(
    XtX = N * R,
    Xty = N * v,
    n = N,
    yty = N
  )
  
  ## 2. Out-of-sample LD
  fit_R0 <- susie_suff_stat(
    XtX = N * R0,
    Xty = N * v,
    n = N,
    yty = N,
    L = 5
  )
  
  ## 3. Naive shrinkage
  lambda <- 1 / (N0 + 1)
  diag_R <- diag(R)
  R_out  <- R0
  k      <- min(N0, J)
  
  love_res <- susie_love_lambda(v, R_out, diag_R,
                                lambda,
                                N, L = 5,
                                k,
                                nu_delta_init = 1000,
                                step_susie = 20,
                                step_R = 30,
                                step_nu = 10,
                                max_iter = 30, 
                                tol = 1e-4)
  
  ## 4. WS shrinkage
  diag_R <- diag(R)
  R_out  <- compute_wen_stephens(R_0 = R0, N_ref = N0)
  k      <- min(N0, J)
  
  love_res_WS <- susie_love_lambda(v, R_out, diag_R,
                                lambda = 0,
                                N, L = 5,
                                k,
                                nu_delta_init = 1000,
                                step_susie = 20,
                                step_R = 30,
                                step_nu = 10,
                                max_iter = 30, 
                                tol = 1e-4)
  
  ## 5. LW shrinkage
  diag_R <- diag(R)
  R_out  <- compute_lw_linear_from_R0(R0, N0)
  k      <- min(N0, J)
  
  love_res_LW_linear <- susie_love_lambda(v, R_out, diag_R,
                                lambda = 0,
                                N, L = 5,
                                k,
                                nu_delta_init = 1000,
                                step_susie = 20,
                                step_R = 30,
                                step_nu = 10,
                                max_iter = 30, 
                                tol = 1e-4)
  
  fitted_rss = list(fit, 
                    fit_R0, 
                    love_res, 
                    love_res_WS, 
                    love_res_LW_linear)
  
  for (method in 1:5){
    if (method < 3) {
      summary_all_CS = summary(fitted_rss[[method]])$cs
      all_CS <- lapply(summary_all_CS$variable, function(x) {
        as.numeric(unlist(strsplit(x, ",")))})
    } else {
      all_CS = get_cs(fitted_rss[[method]]$b_res$alpha, fitted_rss[[method]]$R_q)$cs
    }
    
    L_infer = length(all_CS)
    if (is.null(all_CS) || length(all_CS) == 0) {
      coverages[seed, method] = 0
      powers[seed, method] = 0
      no_CSs[seed, method] = 0
      no_SNPs_CS[seed, method] = 0
    } else{
      num_contains = 0
      all_selected_SNPs = c()
      for (ell in 1:L_infer){
        this_CS = all_CS[[ell]]
        all_selected_SNPs = c(all_selected_SNPs, this_CS)
        num_contains = num_contains + (length(intersect(this_CS, causal_SNPs_loc)) > 0)
      }
      ## coverage = proportion of CS that contains a true casual SNP
      coverages[seed, method] = num_contains / L_infer
      ## power = proportion of casual SNP that is contained in a CS
      selected = length(intersect(all_selected_SNPs, causal_SNPs_loc)) 
      powers[seed, method] = selected / length(causal_SNPs_loc)
      
      ## number of CSs
      no_CSs[seed, method] = L_infer
      
      ## number of SNPs per CSs
      no_SNPs_CS[seed, method] = length(all_selected_SNPs) / L_infer
    }
  }
}


```

# 2. Visualization results

```{r results, fig.width=8, fig.height=6}
library(ggplot2)

list_mat = list(coverages, powers, no_CSs, no_SNPs_CS)
list_name = c("Coverages", "Power", "Number of CSs", "Number of SNPs per CS")
plots = list()
B <- 2000
for (i in 1:4){
  m = list_mat[[i]]
  colnames(m) <- c('In-', 'Out-', 'naive', 'WS', 'LW')
  means <- colMeans(m)

  ci_lower <- numeric(ncol(m))
  ci_upper <- numeric(ncol(m))
  for (j in 1:ncol(m)) {
    boot_means <- replicate(B, mean(sample(m[, j], replace = TRUE)))
    ci_lower[j] <- quantile(boot_means, 0.05)
    ci_upper[j] <- quantile(boot_means, 0.95)
  }

  df <- data.frame(
    variable = factor(colnames(m), levels = colnames(m)),
    mean = means,
    lower = ci_lower,
    upper = ci_upper
  )

  plots[[i]] = ggplot(df, aes(x = variable, y = mean)) +
    geom_point(size = 3, color = "blue") +
    geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
    labs(title = list_name[i],
         x = "Methods", y = "Mean (90% bootstrap CI)")
}

library(patchwork)
wrap_plots(plots, ncol = 2)
```
