---
title: "Test new grid-based optimization for nu_0"
author: "Dat Do"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
date: "2026-06-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 0. Aim
Here we aim to still learn $\nu_0$ using VEB while fixing $\delta$ (i.e., $ = 1$). This is arguably the most principal approach, although the convergence of $\nu_0$ is poorly (it rarely moves away from the intitial point). Therefore, in this workflowr, we seek alternative optimization approach. 

Since optimizing $\nu_0$ is only a 1D optimization problem, I plan to find it using a grid of $\nu_0$. But a grid of 10 values of $\nu_0$ would lead to an optimization problem that is 10 times slower than `susie-love-lambda`. After working with Claude, we decided the best solution is:

1) For each $\nu_0$ in the grid, fix $\nu_0$ (turn `estimate_nu_0` to `FALSE`) and run the optimization of other parameters in 10 iterations. 

2) Choose the best $\nu_0$ and call it the warm start for learning the optimal $\nu_0$

3) Start optimizing everything from this $\nu_0$ (turn `estimate_nu_0` to `TRUE`).

So the total cost is only 10 (grid points) x 10 (iterations) + number of iterations in step 3. Not too bad. Let's see how it work in the low mismatch scenario. 

One more update is we changed the optimization of nu_0 and nu_q to log-scale. It works slightly better. 


## 1. Simulate data from a coalescent model

```{r simulate, fig.width=8, fig.height=6}
library(susielove)

seed  <- 1
T_split <- 25
chrom <- "chr22"
start <- 20e6
end   <- 21e6
N     <- 25000L
N0    <- 200L
J     <- 1000L
h2    <- 0.01

res <- sim_2pop(chrom, start, end,
                T_split = T_split,
                n_gwas = N,
                n_ref = N0,
                J = J,
                h2 = h2,
                seed = seed)

v               <- res$v
R               <- res$R
R0              <- res$R0
beta_true       <- res$beta_true
causal_SNPs_loc <- res$causal_SNPs_loc

plot(v, pch = 16)
points(causal_SNPs_loc, v[causal_SNPs_loc], pch = 16, col = "red")
```

## 2. Try fitting susieR with in-sample and out-of-sample covariance matrices

```{r susie-insample, fig.width=8, fig.height=6}
library(susieR)

fit <- susie_suff_stat(
  XtX = N * R,
  Xty = N * v,
  n = N,
  yty = N
)
susie_plot(fit, y = "PIP", b = beta_true, main = "In-sample cov mat")
```

```{r susie-outsample, fig.width=8, fig.height=6}
library(susieR)

fit_R0 <- susie_suff_stat(
  XtX = N * R0,
  Xty = N * v,
  n = N,
  yty = N
)
susie_plot(fit_R0, y = "PIP", b = beta_true, main = "Out-of-sample cov mat")
```


```{r susie-outsample-diag, fig.width=8, fig.height=6}
d0_inv_sqrt <- 1 / sqrt(diag(R0))
LD0 <- diag(d0_inv_sqrt) %*% R0 %*% diag(d0_inv_sqrt)
d_sqrt <- sqrt(diag(R))
R_tilde <- diag(d_sqrt) %*% LD0 %*% diag(d_sqrt)

fit <- susie_suff_stat(
  XtX = N * R_tilde,
  Xty = N * v,
  n = N,
  yty = N
)
susie_plot(fit, y = "PIP", b = beta_true, main = "Out-of-sample cov mat (corrected diagonal)")
```


## 3. Fit susie_XL with new log-space optimization (nu_0_init = 1000)

```{r prep-inputs}
d0_inv_sqrt <- 1 / sqrt(diag(R0))
LD0 <- diag(d0_inv_sqrt) %*% R0 %*% diag(d0_inv_sqrt)
d_sqrt <- sqrt(diag(R))
R_tilde <- diag(d_sqrt) %*% LD0 %*% diag(d_sqrt)

diag_R <- diag(R)
R_out  <- R_tilde
k      <- min(N0, J)
delta  <- 1
```

```{r susie-XL-1000, fig.width=8, fig.height=6}
nu_0_init <- 1000
max_iter  <- 100

XL_res_1000 <- susie_XL(v, R_out, diag_R,
                         delta,
                         N, L = 5,
                         k,
                         nu_0_init     = nu_0_init,
                         max_iter_init = 100,
                         step_susie    = 20,
                         step_R        = 100,
                         estimate_nu_0 = TRUE,
                         estimate_delta = FALSE,
                         max_iter      = max_iter,
                         tol           = 1e-4)

b_res <- XL_res_1000$b_res
R_q   <- XL_res_1000$R_q

susie_plot_alpha(
  alpha = b_res$alpha,
  R     = cov2cor(R_q),
  b     = beta_true,
  main  = paste0("SuSiE-XL (nu_0_init=1000) | nu_0=",
                 round(XL_res_1000$nu_0, 1),
                 " | nu_q=", round(XL_res_1000$nu_q, 1))
)
```

## 4. Fit susie_XL with new log-space optimization (nu_0_init = 100)

```{r susie-XL-100, fig.width=8, fig.height=6}
nu_0_init <- 100
max_iter  <- 100

XL_res_100 <- susie_XL(v, R_out, diag_R,
                        delta,
                        N, L = 5,
                        k,
                        nu_0_init     = nu_0_init,
                        max_iter_init = 500,
                        step_susie    = 20,
                        step_R        = 100,
                        estimate_nu_0 = TRUE,
                        estimate_delta = FALSE,
                        max_iter      = max_iter,
                        tol           = 1e-4)

b_res <- XL_res_100$b_res
R_q   <- XL_res_100$R_q

susie_plot_alpha(
  alpha = b_res$alpha,
  R     = cov2cor(R_q),
  b     = beta_true,
  main  = paste0("SuSiE-XL (nu_0_init=100) | nu_0=",
                 round(XL_res_100$nu_0, 1),
                 " | nu_q=", round(XL_res_100$nu_q, 1))
)
```

## 5. Compare convergence

Check whether the two initializations now converge to the same nu_0 and nu_q.

```{r compare}
cat("nu_0_init = 1000 -> nu_0:", round(XL_res_1000$nu_0, 2),
    " | nu_q:", round(XL_res_1000$nu_q, 2),
    " | iters:", length(XL_res_1000$loss_history), "\n")
cat("nu_0_init = 100  -> nu_0:", round(XL_res_100$nu_0, 2),
    " | nu_q:", round(XL_res_100$nu_q, 2),
    " | iters:", length(XL_res_100$loss_history), "\n")

plot(XL_res_100$loss_history, type = "l", col = "blue",
     xlab = "Outer iteration", ylab = "Loss", main = "Convergence (loss)")
lines(XL_res_1000$loss_history, col = "red")
legend("topright", legend = c("nu_0_init=100", "nu_0_init=1000"),
       col = c("blue", "red"), lty = 1)
```

**Conclusion:** Optimal $\nu_0$ highly depends on the initialization.

## 6. Profile empirical Bayes over nu_0

The log-space `optimize()` finds the right nu_0 for a fixed R_q, but R_q itself was
optimized for the current nu_0 — so the two steps keep each other near the
initialization (a CAVI fixed-point coupling issue). The fix is a two-phase
profile search: short runs across a log-spaced grid to rank candidates cheaply,
then one full run warm-started from the best.

### Phase 1: coarse scan (10 iterations each)

```{r profile-coarse}
nu_0_grid <- exp(seq(log(10), log(5*N0), length.out = 10))

coarse_results <- lapply(nu_0_grid, function(nu_0_fixed) {
  susie_XL(v, R_out, diag_R, delta, N, L = 5, k,
            nu_0_init     = nu_0_fixed,
            estimate_nu_0 = FALSE,
            max_iter_init = 100,
            step_susie    = 10,
            step_R        = 20,
            max_iter      = 10,
            tol           = -Inf)
})

coarse_loss <- sapply(coarse_results, function(r) tail(r$loss_history, 1))
best_idx    <- which.min(coarse_loss)

plot(nu_0_grid, coarse_loss, log = "x", type = "b", pch = 16,
     xlab = "nu_0 (log scale)", ylab = "Loss after 10 iters",
     main = "Profile loss — coarse scan")
abline(v = nu_0_grid[best_idx], col = "red", lty = 2)
cat("Best nu_0 on grid:", round(nu_0_grid[best_idx], 2),
    "| coarse loss:", round(coarse_loss[best_idx], 4), "\n")
```

### Phase 2: full convergence from the best warm start

```{r profile-full, fig.width=8, fig.height=6}
best_res <- susie_XL(v, R_out, diag_R, delta, N, L = 5, k,
                      nu_0_init     = nu_0_grid[best_idx],
                      z_init        = coarse_results[[best_idx]]$Z_final,
                      estimate_nu_0 = TRUE,
                      max_iter_init = 100,
                      step_susie    = 10,
                      n_inner = 5,
                      step_R        = 50,
                      max_iter      = 100,
                      tol           = 1e-4)

cat("Converged loss:", round(tail(best_res$loss_history, 1), 4),
    "| nu_0:", round(best_res$nu_0, 2),
    "| nu_q:", round(best_res$nu_q, 2),
    "| iters:", length(best_res$loss_history), "\n")

susie_plot_alpha(
  alpha = best_res$b_res$alpha,
  R     = cov2cor(best_res$R_q),
  b     = beta_true,
  main  = paste0("SuSiE-XL profile EB | nu_0=", round(best_res$nu_0, 1),
                 " | nu_q=", round(best_res$nu_q, 1))
)
```

**Observation:** With a good starting $\nu_0$, the algorithm often converges within a few iterations.


## 7. Profile EB with warm-starts across grid points (slightly smarter solution)

In section 6 the 10-iteration budget is spent mostly on the initial transient
because each grid point starts cold. Here we chain the grid runs: each point
is warm-started with `Z_final` from its left neighbour, so the transient is
already past and the 10 iterations reflect the actual profile loss.

The chosen $log(\nu_0)$ here can be slightly different from section 6.

### Phase 1: warm-started coarse scan

```{r profile-warm-coarse}
nu_0_grid_w <- exp(seq(log(10), log(5*N0), length.out = 10))

coarse_results_w <- vector("list", length(nu_0_grid_w))
z_prev <- NULL

for (i in seq_along(nu_0_grid_w)) {
  coarse_results_w[[i]] <- susie_XL(v, R_out, diag_R, delta, N, L = 5, k,
                                     nu_0_init     = nu_0_grid_w[i],
                                     z_init        = z_prev,
                                     estimate_nu_0 = FALSE,
                                     max_iter_init = 100,
                                     step_susie    = 10,
                                     step_R        = 20,
                                     max_iter      = 10,
                                     tol           = -Inf)
  z_prev <- coarse_results_w[[i]]$Z_final
}

coarse_loss_w <- sapply(coarse_results_w, function(r) tail(r$loss_history, 1))
best_idx_w    <- which.min(coarse_loss_w)

plot(nu_0_grid_w, coarse_loss_w, log = "x", type = "b", pch = 16,
     xlab = "nu_0 (log scale)", ylab = "Loss after 10 iters",
     main = "Profile loss — warm-started coarse scan")
abline(v = nu_0_grid_w[best_idx_w], col = "red", lty = 2)
cat("Best nu_0 on grid:", round(nu_0_grid_w[best_idx_w], 2),
    "| coarse loss:", round(coarse_loss_w[best_idx_w], 4), "\n")
```

### Phase 2: full convergence from the best warm start

```{r profile-warm-full, fig.width=8, fig.height=6}
best_res_w <- susie_XL(v, R_out, diag_R, delta, N, L = 5, k,
                        nu_0_init     = nu_0_grid_w[best_idx_w],
                        z_init        = coarse_results_w[[best_idx_w]]$Z_final,
                        estimate_nu_0 = TRUE,
                        max_iter_init = 100,
                        step_susie    = 20,
                        step_R        = 50,
                        max_iter      = 100,
                        tol           = 1e-4)

cat("Converged loss:", round(tail(best_res_w$loss_history, 1), 4),
    "| nu_0:", round(best_res_w$nu_0, 2),
    "| nu_q:", round(best_res_w$nu_q, 2),
    "| iters:", length(best_res_w$loss_history), "\n")

susie_plot_alpha(
  alpha = best_res_w$b_res$alpha,
  R     = cov2cor(best_res_w$R_q),
  b     = beta_true,
  main  = paste0("SuSiE-XL warm profile EB | nu_0=", round(best_res_w$nu_0, 1),
                 " | nu_q=", round(best_res_w$nu_q, 1))
)
```
