---
title: "Various shrinkage estimates for the out-of-sample correlation matrix"
author: "dodat-stats"
date: "2026-06-01"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## 1. Model and shrinkage estimate of the out-of-sample population correlation matrix

Let's suppose that we want to go with the `susie_love_lambda` implementation of the method. So that we assume the prior
$$R | \nu_{\delta} \sim \mathcal{IW}(\nu_{\delta} \Psi_0, \nu_{\delta} + J + 1), $$
where $\Psi_0$ is somewhat shrinkage estimate obtained from the out-of-sample $R_0$. We can think of $\Psi_0$ as an estimate for the population out-of-sample covariance matrix. For the uniform presentation, we consider CORRELATION MATRIX instead of covariance here (so that the diagonal is constrained to be 1). Consider the following strategies:

1. Naive shrinkage: $\Psi_{naive} = \left(1-\dfrac{1}{N_0}) R_0 + \dfrac{1}{N_0} I$

2. Wen and Stephens (2010) shrinkage: $\Psi_{WS} = \left(1- \lambda)^2 R_0 + (1 - (1-\lambda)^2) I$, where $\lambda = \dfrac{\theta}{\theta + 2 N_0}$ and $\theta$ is the Watterson's estimator $\theta = \left(\sum_{i=1}^{2N_0 - 1} \dfrac{1}{i}\right)^{-1} \approx (\log(2N_0) + 0.57)^{-1} $.

3. Ledoit and Wolf (2004) linear shrinkage: $\Psi_{linear-LW} = \left(1- \lambda) R_0 + \lambda I$, where $\lambda \asymp \dfrac{1}{N_0}$. 

4. Ledoit and Wolf (2020) non-linear shrinkage: $\Psi_{nonlinear-LW}$ is obtained from non-linear shrinkage of the sample eigenvalues. See their AoS 2020 paper, which use the random matrix theory to derive the shrinakge rule. This is considered state-of-the-art method for shrinkage in their review paper later.

Both 3 and 4 are implemented in the package `cvCovEst`.

```{r}


compute_wen_stephens <- function(R_0, N_ref) {
  # Number of haplotypes (diploid population)
  d <- 2 * N_ref
  
  # Calculate Watterson's estimator constant (Harmonic sum)
  harmonic_sum <- sum(1 / (1:(d - 1)))
  theta        <- 1 / harmonic_sum
  
  # Calculate the exact mutation shrinkage parameter lambda
  lambda <- theta / (theta + 2 * N_ref)
  
  # Apply local uniform shrinkage: Psi = (1-lambda)^2 * R_0 + (1 - (1-lambda)^2) * I
  p <- ncol(R_0)
  R_ws <- (1 - lambda)^2 * R_0 + (1 - (1 - lambda)^2) * diag(p)
  
  return(R_ws)
}


compute_lw_linear_from_R0 <- function(R_0, N_ref) {
  p <- ncol(R_0)
  
  # 1. Sum of the squared off-diagonal elements of R_0
  # (This is the Frobenius norm of the off-diagonals)
  sum_sq_off_diag <- sum(R_0^2) - p
  
  # 2. Estimate the sampling variance (noise) parameter under normality
  # In correlation space, the sum of variances of r_ij is approximated by:
  beta_sq <- (1 / N_ref) * (p + sum_sq_off_diag)
  
  # 3. Calculate the distance to the identity target matrix
  # Since diag(R_0) is all 1s and diag(I) is all 1s, the distance is just 
  # the sum of the squared off-diagonal elements.
  delta_sq <- sum_sq_off_diag
  
  # 4. Compute the optimal shrinkage intensity alpha
  # Guard against division by zero if R_0 is already the identity matrix
  if (delta_sq == 0) {
    alpha <- 1
  } else {
    alpha <- min(1, beta_sq / delta_sq)
  }
  
  # 5. Construct the shrunk matrix: alpha * I + (1 - alpha) * R_0
  R_lw_linear <- (1 - alpha) * R_0 + alpha * diag(p)
  
  # Print the learned alpha for transparency
  cat("Ledoit-Wolf Linear Alpha estimated from R_0:", alpha, "\n")
  
  return(R_lw_linear)
}

compute_lw_nonlinear_from_R0 <- function(R_0, N_ref) {
  p <- ncol(R_0)
  
  # 1. Spectral Decomposition of the correlation matrix
  decomp <- eigen(R_0, symmetric = TRUE)
  lambda <- decomp$values    # Sample eigenvalues
  U      <- decomp$vectors   # Sample eigenvectors
  
  # Enforce non-negativity (numerical cleanup for trailing eigenvalues)
  lambda <- pmax(lambda, 0)
  
  # 2. Setup the non-linear shrinkage parameters (Random Matrix Theory)
  gamma <- p / N_ref
  
  # Compute the complex grid parameters safely
  h <- N_ref^(-1/3)
  z <- lambda + 1i * h  # z is a vector of length p
  
  # 3. Calculate the Stieltjes transform (m)
  # Instead of duplicating matrices, we use outer subtraction:
  # column j evaluates 1 / (z_j - lambda_i)
  # rowMeans then averages across all i for each j.
  m <- rowMeans(1 / (outer(z, lambda, "-")))
  
  # 4. Calculate the optimal non-linear shrinkage function (d_j)
  d <- lambda / ((1 - gamma - gamma * lambda * Re(m))^2 + (gamma * lambda * Im(m))^2)
  
  # Floor the cleaned eigenvalues to guarantee strict positive-definiteness
  d <- pmax(d, min(lambda[lambda > 0]))
  
  # 5. Reconstruct the regularized matrix using the cleaned eigenvalues
  R_lw_nonlinear_raw <- U %*% (d * t(U))
  
  # 6. Standardize back to a perfect correlation space (diagonals = 1)
  R_lw_nonlinear <- cov2cor(R_lw_nonlinear_raw)
  
  return(R_lw_nonlinear)
}

```


# 1. Preparation

## 1.1. Simulate data from a coalescent model

First we use `sim_2pop()` to simulate two populations that diverged 25 generations
ago. POP1 is a large GWAS cohort (N = 25,000) and POP2 is a small LD
reference panel (N₀ = 200).

```{r simulate, fig.width=8, fig.height=6}
library(susielove)

seed  <- 1
T_split <- 25
chrom <- "chr22"
start <- 20e6
end   <- 21e6
N     <- 25000L
N0    <- 200L
J     <- 1000L
h2    <- 0.01

res <- sim_2pop(chrom, start, end,
                T_split = T_split,
                n_gwas = N,
                n_ref = N0,
                J = J,
                h2 = h2,
                seed = seed)

v               <- res$v
R               <- res$R
R0              <- res$R0
beta_true       <- res$beta_true
causal_SNPs_loc <- res$causal_SNPs_loc

```


Here are marginal associations:
```{r}

## standardized
d = 1 / sqrt(diag(R))
v = d * v
R = cov2cor(R)
R0 = cov2cor(R0)

plot(v, pch = 16)
points(causal_SNPs_loc, v[causal_SNPs_loc], pch = 16, col = "red")

```

## 1.2. Fit SuSiE with in-sample LD matrix (oracle baseline)

When the true in-sample LD is available, standard SuSiE works perfectly:

```{r susie-insample, fig.width=8, fig.height=6}
library(susieR)

fit <- susie_suff_stat(
  XtX = N * R,
  Xty = N * v,
  n = N,
  yty = N
)
susie_plot(fit, y = "PIP", b = beta_true, main = "In-sample LD")
```

## 1.3. Fit SuSiE with out-of-sample LD

Using the mismatched LD reference directly leads to spurious signals:

```{r susie-outsample, fig.width=8, fig.height=6}
fit_R0 <- susie_suff_stat(
  XtX = N * R0,
  Xty = N * v,
  n = N,
  yty = N,
  L = 5
)
susie_plot(fit_R0, y = "PIP", b = beta_true, main = "Out-of-sample LD mat")
```

# 2. Fit SuSiE-LovE

## 2.1. naive shrinkage

`susie_love_lambda` jointly learns the covariance matrix while performing
fine-mapping. Here $\lambda$ is set to $1 / (N_0 + 1) \approx 1 / 200 = 0.005$

```{r susie-love-naive-shrink, fig.width=8, fig.height=6}

lambda <- 1 / (N0 + 1)
diag_R <- diag(R)
R_out  <- R0
k      <- min(N0, J)

love_res <- susie_love_lambda(v, R_out, diag_R,
                              lambda,
                              N, L = 5,
                              k,
                              nu_delta_init = 1000,
                              step_susie = 20,
                              step_R = 30,
                              step_nu = 10,
                              max_iter = 30, 
                              tol = 1e-4)

b_res <- love_res$b_res
R_q   <- love_res$R_q

susie_plot_alpha(
  alpha = b_res$alpha,
  R = cov2cor(R_q),
  b = beta_true,
  main = paste0("SuSiE-LovE with lambda = ", lambda)
)
```
The result looks good. The spurious CS is gone. 

## 2.2. Wen and Stephens (WS) shrinkage

Here (and other 2 methods) I directly obtain the estimated shrinkage $\Psi$ and set $\lambda$ in the algorithm to 0.

```{r susie-love-WS, fig.width=8, fig.height=6}

diag_R <- diag(R)
R_out  <- compute_wen_stephens(R_0 = R0, N_ref = N0)
k      <- min(N0, J)

love_res_WS <- susie_love_lambda(v, R_out, diag_R,
                              lambda = 0,
                              N, L = 5,
                              k,
                              nu_delta_init = 1000,
                              step_susie = 20,
                              step_R = 30,
                              step_nu = 10,
                              max_iter = 30, 
                              tol = 1e-4)

b_res <- love_res_WS$b_res
R_q   <- love_res_WS$R_q

susie_plot_alpha(
  alpha = b_res$alpha,
  R = cov2cor(R_q),
  b = beta_true,
  main = paste0("SuSiE-LovE with WS shrinakge")
)
```




## 2.3. LW linear shrinkage

```{r susie-love-linear-LW, fig.width=8, fig.height=6}

diag_R <- diag(R)
R_out  <- compute_lw_linear_from_R0(R0, N0)
k      <- min(N0, J)

love_res_LW_linear <- susie_love_lambda(v, R_out, diag_R,
                              lambda = 0,
                              N, L = 5,
                              k,
                              nu_delta_init = 1000,
                              step_susie = 20,
                              step_R = 30,
                              step_nu = 10,
                              max_iter = 30, 
                              tol = 1e-4)

b_res <- love_res_LW_linear$b_res
R_q   <- love_res_LW_linear$R_q

susie_plot_alpha(
  alpha = b_res$alpha,
  R = cov2cor(R_q),
  b = beta_true,
  main = paste0("SuSiE-LovE with linear LW shinkage")
)
```



## 2.4. LW non-linear shrinkage

```{r susie-love-nonlinear-LW, fig.width=8, fig.height=6}

diag_R <- diag(R)
R_out  <- compute_lw_nonlinear_from_R0(R0, N0)
k      <- min(N0, J)

love_res_LW_nonlinear <- susie_love_lambda(v, R_out, diag_R,
                              lambda = 0,
                              N, L = 5,
                              k,
                              nu_delta_init = 1000,
                              step_susie = 20,
                              step_R = 30,
                              step_nu = 10,
                              max_iter = 30, 
                              tol = 1e-4)

b_res <- love_res_LW_nonlinear$b_res
R_q   <- love_res_LW_nonlinear$R_q

susie_plot_alpha(
  alpha = b_res$alpha,
  R = cov2cor(R_q),
  b = beta_true,
  main = paste0("SuSiE-LovE with non-linear LW shrinakge")
)
```

It is a bit problematic for this shrinkage since many eigenvalues are still close to 0. Our theory for conservative behavior requires the eigenvalues of $\Psi$ to be bounded away from 0 to work (Lower bound of the `L_cov` in the $L > 2$ case; I can show exactly where it is needed when we meet).

```{r}
eigen(R_out)$values
```

Adding a small ridge will help.
```{r susie-love-nonlinear-LW-modify-little, fig.width=8, fig.height=6}

diag_R <- diag(R)
R_out  <- compute_lw_nonlinear_from_R0(R0, N0)
k      <- min(N0, J)

love_res_LW_nonlinear <- susie_love_lambda(v, R_out, diag_R,
                              lambda = 1e-3,
                              N, L = 5,
                              k,
                              nu_delta_init = 1000,
                              step_susie = 20,
                              step_R = 30,
                              step_nu = 10,
                              max_iter = 30, 
                              tol = 1e-4)

b_res <- love_res_LW_nonlinear$b_res
R_q   <- love_res_LW_nonlinear$R_q

susie_plot_alpha(
  alpha = b_res$alpha,
  R = cov2cor(R_q),
  b = beta_true,
  main = paste0("SuSiE-LovE with non-linear LW shrinakge")
)
```

# 3. Conclusion

I think I will go with the naive, WS, and LW linear shrinkage now. So we have automatic choice for $\lambda$. Learning $\lambda$ directly from the susie_love model is problematic due to the large dimension $J$ and that we are considering IW distribution (so it prefers well-condition prior mean and set $\lambda$ to large). Having automatic choice of $\lambda$ is great, especially when it is motivated from well-established framework like Wen and Stephens (2010) or Ledoit and Wolf (2004). 


